# Analysis - Microenvironment, post DESeq

Set libpaths

```{r}
# Set libpaths
.libPaths(c("/raleighlab/data1/czou/gbm_perturb/gbm_perturb/renv/library/R-4.3/x86_64-pc-linux-gnu", .libPaths()))
```

Import libraries

```{r}
library(Seurat)
library(clusterProfiler)
library(org.Hs.eg.db)
library(knitr)
library(msigdbr)
library(fgsea)
library(dplyr)
library(data.table)
library(ggplot2)
library(grid)
library(ComplexHeatmap)
library(colorRamp2)
library(tidyr)
library(patchwork)
library(ggtree)
library(gridExtra)
library(ape)
library(RColorBrewer)
library(reactome.db)
library(annotables)
library(biomaRt)
```

## Validation of non-targeting guides

Define constants

```{r}
CELL.TYPE <- commandArgs(trailingOnly=TRUE)[1]
GENE.AGGREGATION <- "sgRNA_gene_sgNegCtrl3"
INPUT.DIR <- sprintf("/raleighlab/data1/czou/gbm_perturb/gbm_perturb_sb28_clean_outputs/deseq/SB28_integrated_20231207_ME/%s/%s", GENE.AGGREGATION, CELL.TYPE)
OUTPUT.DIR <- sprintf("/raleighlab/data1/czou/gbm_perturb/gbm_perturb_sb28_clean_outputs/gsea/%s/%s", GENE.AGGREGATION, CELL.TYPE)
```

Import the DESeq output

```{r}
data.list.symbols <- list()
for (dir in INPUT.DIR) {
  file_names <- list.files(dir)
  for (i in file_names) {
    name <- gsub(".csv", "", i)
    df <- read.table(paste(dir, "/", i, sep = ""))
    data.list.symbols[[name]] <- df
  }
}
```

Take a look at what these outputs look like after going through and filtering for validity

```{r}
processed.deseq.tables <- list()
for (name in names(data.list.symbols)) {
  deseq_table <- data.list.symbols[[name]]
  deseq_table <- deseq_table[!is.na(deseq_table$pvalue), ]
  deseq_table <- deseq_table[!is.na(deseq_table$padj), ]
  processed.deseq.tables[[name]] <- deseq_table
}
data.list <- processed.deseq.tables
```

For each combination, get the number of DE genes.

```{r}
df <- data.frame(matrix(nrow = length(data.list), ncol = 4, dimnames = list(names(data.list), c("Comparison", "n_upregulated_genes", "n_downregulated_genes", "n_total"))))
for (name in names(data.list)) {
  deseq_table <- data.list[[name]]
  upregulated.table <- deseq_table[deseq_table$log_fc > 0.5, ]
  downregulated.table <- deseq_table[deseq_table$log_fc < -0.5, ]
  df[name, "n_upregulated_genes"] <- dim(upregulated.table)[1]
  df[name, "n_downregulated_genes"] <- dim(downregulated.table)[1]
  df[name, "n_total"] <- dim(deseq_table)[1]
  df[name, "Comparison"] <- name
}
```

Plot

```{r}
# Reshape the data from wide to long format
df_long <- pivot_longer(df, cols = c(n_upregulated_genes, n_downregulated_genes, n_total),
                        names_to = "Category", values_to = "Count")

# Create the bar plot
plot <- ggplot(df_long, aes(x = Comparison, y = Count, fill = Category)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  scale_fill_brewer(palette = "Set1") +  # You can choose a palette that suits your data
  theme_minimal() +
  labs(x = "Comparison", y = "Count", fill = "Category") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
ggsave(paste(OUTPUT.DIR, "comparison_of_non-targeting_degenes.png", sep = "/"), plot, height = 12, width = 15)
```

## Individual cell type GSEA

Define constants

```{r}
CELL.TYPE <- "OPCs"
GENE.AGGREGATION <- "sgRNA_gene_sgNegCtrl3_knockouts"
INPUT.DIR <- sprintf("/raleighlab/data1/czou/gbm_perturb/gbm_perturb_sb28_clean_outputs/deseq/SB28_integrated_20231207_ME/%s/%s", GENE.AGGREGATION, CELL.TYPE)
OUTPUT.DIR <- sprintf("/raleighlab/data1/czou/gbm_perturb/gbm_perturb_sb28_clean_outputs/gsea/%s/%s_pval", GENE.AGGREGATION, CELL.TYPE)
```

Import the DESeq output

```{r}
data.list.symbols <- list()
for (dir in INPUT.DIR) {
  file_names <- list.files(dir)
  for (i in file_names) {
    name <- gsub(".csv", "", i)
    df <- read.table(paste(dir, "/", i, sep = ""))
    data.list.symbols[[name]] <- df
  }
}
```

Prepare an EMSEMBL gene ID list

```{r}
mapping.df <- annotables::grcm38
missing.genes <- list()
data.list.ensembl <- list()
for (name in names(data.list.symbols)) {
  deseq_table <- data.list.symbols[[name]]
  mapping.df <- mapping.df[!duplicated(mapping.df$symbol), ]
  mapping.table <- mapping.df[mapping.df$symbol %in% deseq_table$feature, ]
  mapping.table <- as.data.frame(mapping.table)
  rownames(mapping.table) <- mapping.table$symbol
  deseq_table$ensembl <- mapping.table[deseq_table$feature, "ensgene"]
  deseq_table.missing <- deseq_table[is.na(deseq_table$ensembl), ]
  missing.genes[[name]] <- deseq_table.missing$feature
  data.list.ensembl[[name]] <- deseq_table[!is.na(deseq_table$ensembl), ]
}
write.table(missing.genes[[1]], file = paste(OUTPUT.DIR, "missing_genes.txt", sep = "/"), row.names = FALSE, quote = FALSE)
```

Get rid of anything that shows low expression:

```{r}
processed.deseq.tables <- list()
for (name in names(data.list.ensembl)) {
  deseq_table <- data.list.ensembl[[name]]
  deseq_table <- deseq_table[!is.na(deseq_table$pvalue), ]
  deseq_table <- deseq_table[!is.na(deseq_table$padj), ]
  deseq_table <- deseq_table[deseq_table$pvalue < 0.05, ]
  processed.deseq.tables[[name]] <- deseq_table
}
data.list <- processed.deseq.tables
```

Create an ordered list from the processed DESeq tables:

```{r}
ranked_gene_set_list <- list()
for (name in names(processed.deseq.tables)) {
  deseq_table <- processed.deseq.tables[[name]]
  # Group by unique Ensembl ID and average
  deseq_table <- deseq_table %>%
    group_by(ensembl) %>%
    summarize(log_fc = mean(log_fc))
  deseq_table <- deseq_table %>% mutate(rank = rank(log_fc,  ties.method = "random"))
  deseq_table <- deseq_table[order(-deseq_table$rank),] # Rank deterministically
  
  # Generate a ranked list
  gene_list <- deseq_table$log_fc
  gene_names <- deseq_table$ensembl
  names(gene_list) <- gene_names
  ranked_gene_set_list[[name]] <- gene_list
}
```

Run GSEA using `msigdbr` and `fgsea`.

```{r}
msigdbr_df <- msigdbr(species = "Mus musculus", category = "H")
msigdbr_list = split(x = msigdbr_df$ensembl_gene, f = msigdbr_df$gs_name)

set.seed(1) # required for deterministic performance
fgsea.output.list <- list()
for (name in names(ranked_gene_set_list)) {
  gsea_results <- fgsea(pathways = msigdbr_list, stats = ranked_gene_set_list[[name]],
                        maxSize = 500, eps = 0.0) # allow arbitrarily low p-values
  fgsea.output.list[[name]] <- gsea_results
}
saveRDS(fgsea.output.list, paste(OUTPUT.DIR, "fgsea_ensembl_ontology_list.rds", sep = "/"))
```

### Heatmaps, clustered and alphabetical

```{r}
extracted_cols <- lapply(names(fgsea.output.list), function(name) {
  col <- fgsea.output.list[[name]]$ES
  names(col) <- fgsea.output.list[[name]]$pathway
  col
})
combined.df <- do.call(cbind, extracted_cols)
colnames(combined.df) <- names(fgsea.output.list)

ht = Heatmap(combined.df,
        name = "Enrichment Scores across Perturbations and Conditions",
        col = colorRamp2(c(-1, 0, 1), c("blue", "white", "red")),
        cluster_column_slices = FALSE,
        cluster_columns = FALSE,
        width = ncol(combined.df)*unit(4, "mm"),
        height = nrow(combined.df)*unit(4, "mm"),
)
pdf(paste(OUTPUT.DIR, "enrichment_combined_matrix_alphabetical.pdf", sep = "/"),
    width=35,
    height=19,
    useDingbats=FALSE)
draw(ht, heatmap_legend_side = "top", annotation_legend_side = "top")
dev.off()

extracted_cols <- lapply(names(fgsea.output.list), function(name) {
  col <- fgsea.output.list[[name]]$ES
  names(col) <- fgsea.output.list[[name]]$pathway
  col
})
combined.df <- do.call(cbind, extracted_cols)
colnames(combined.df) <- names(fgsea.output.list)

ht = Heatmap(combined.df,
        name = "Enrichment Scores across Perturbations and Conditions",
        col = colorRamp2(c(-1, 0, 1), c("blue", "white", "red")),
        cluster_column_slices = FALSE,
        cluster_columns = TRUE,
        width = ncol(combined.df)*unit(4, "mm"),
        height = nrow(combined.df)*unit(4, "mm"),
)
pdf(paste(OUTPUT.DIR, "enrichment_combined_matrix_clustered.pdf", sep = "/"),
    width=35,
    height=19,
    useDingbats=FALSE)
draw(ht, heatmap_legend_side = "top", annotation_legend_side = "top")
dev.off()
```

### Bubble plot, clustered

```{r}
combined.df <- bind_rows(fgsea.output.list, .id = "Perturbations")
combined.df$pathway <- gsub("HALLMARK_", "", combined.df$pathway)
combined.df$pathway <- gsub(" ", "", combined.df$pathway)
combined.df$log10pvalue <- -log10(combined.df$padj)
combined.df$Perturbations <- gsub("_non-targeting_noRT", "", combined.df$Perturbations)
unfiltered.df <- combined.df

# combined.df <- combined.df %>%
#   group_by(pathway) %>%
#   filter(any(padj < 0.05)) %>%
#   ungroup()

combined.df <- combined.df %>%
  group_by(pathway) %>%
  filter(any(pval < 0.05)) %>%
  ungroup()

es.df <- combined.df %>%
  dplyr::select(-pval, -padj, -log2err, -NES, -size, -leadingEdge, -log10pvalue)

es.df <- es.df %>%
  group_by(pathway, Perturbations) %>%
  pivot_wider(names_from = Perturbations, values_from = ES) %>%
  group_by(pathway)

es.df <- as.data.frame(es.df)
rownames(es.df) <- es.df$pathway
es.df <- es.df %>% dplyr::select(-pathway)

row_dist <- dist(es.df, method = "euclidean")
row_hclust <- hclust(row_dist, method = "complete")
row_order <- order.dendrogram(as.dendrogram(row_hclust))

col_dist <- dist(t(es.df), method = "euclidean")
col_hclust <- hclust(col_dist, method = "complete")
col_order <- order.dendrogram(as.dendrogram(col_hclust))

df.ordered <- combined.df %>%
  arrange(match(pathway, rownames(es.df)[row_order]), match(Perturbations, colnames(es.df)[col_order]))

df.ordered$pathway <- factor(df.ordered$pathway, levels = rownames(es.df)[row_order])
df.ordered$Perturbations <- factor(df.ordered$Perturbations, levels = colnames(es.df)[col_order])

# Add significance layer
df.ordered$Significance <- ifelse(df.ordered$padj < 0.05, "adjp < 0.05", "NA")

# Clean up names
df.ordered$Pathways <- df.ordered$pathway
df.ordered$Pathways <- gsub("_", " ", df.ordered$Pathways)
```

Make the plot

```{r}
bubble.plot <- ggplot(df.ordered,
               aes(x = Perturbations)) +
  
  # Add the bubble and outline significance layers
  geom_point(aes(y = Pathways, size = log10pvalue, color = ES), alpha = 1.0) +
  geom_point(data = subset(df.ordered, padj < 0.05),
            aes(y = Pathways, size = log10pvalue, shape = 'adjp < 0.05'),
            color = "black", fill = NA, alpha = 0.8) +
  geom_point(data = subset(df.ordered, is.na(pval) | is.na(padj) | is.na(log10pvalue)), 
            aes(y = Pathways), 
            color = "black", size = 1, alpha = 0.6) +
  scale_size(range = c(2, 7)) +
  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "grey50"
  ) +
  scale_shape_manual(name = "Significance",
                   values = c(`adjp < 0.05` = 21),
                   guide = guide_legend(override.aes = list(color = "black", fill = NA))) +

  # Labels
  labs(
    color = "Effect size",
    size = "-Log10 padj"
  ) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.key.height = unit(0.3, 'cm'),
        axis.line = element_line(colour = "black"),
        legend.key = element_blank())

annotation.plot.theme <- theme(
  panel.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  axis.text.x = element_blank(),
  axis.title.x = element_blank(),
  axis.ticks.x = element_blank(),
  axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0))
)

perturb.dendrogram <- ggtree(as.phylo(col_hclust)) + theme_tree() + coord_flip() + scale_x_reverse()
module.dendrogram <- ggtree(as.phylo(row_hclust)) + theme_tree()

vertical.plot <- perturb.dendrogram / bubble.plot + plot_layout(heights = c(0.5, 3))
dendrogram.vertical.plot <- plot_spacer() / module.dendrogram + plot_layout(heights = c(0.17, 1))
combined.plot <- dendrogram.vertical.plot | vertical.plot
combined.plot <- combined.plot + plot_layout(widths = c(0.2, 0.8)) + plot_annotation(
  title = "Enrichment of Hallmark pathways across perturbations",
  subtitle = paste(CELL.TYPE, GENE.AGGREGATION)
)
ggsave(paste(OUTPUT.DIR, "enrichment_bubble_plot_fgsea_clustered.pdf", sep = "/"),
       combined.plot, height = 6, width = 6, device = pdf)
```

Save the aesthetic settings we're using across cells and across the gene aggregate/not

```{r}
# Aesthetics settings for sgRNA_gene
# perturb.dendrogram <- ggtree(as.phylo(col_hclust)) + theme_tree() + coord_flip() + scale_x_reverse()
# module.dendrogram <- ggtree(as.phylo(row_hclust)) + theme_tree()
# 
# vertical.plot <- perturb.dendrogram / bubble.plot + plot_layout(heights = c(0.5, 3))
# dendrogram.vertical.plot <- plot_spacer() / module.dendrogram + plot_layout(heights = c(0.17, 1))
# combined.plot <- dendrogram.vertical.plot | vertical.plot
# combined.plot <- combined.plot + plot_layout(widths = c(0.2, 0.8)) + plot_annotation(
#   title = "Enrichment of Hallmark pathways across perturbations",
#   subtitle = paste(CONTEXT, CONDITION, "noRTNormalized")
# )
# ggsave(paste(OUTPUT.DIR, "enrichment_bubble_plot_fgsea_2.pdf", sep = "/"),
#        combined.plot, height = 6, width = 6, device = pdf)

# vertical.plot <- perturb.dendrogram / bubble.plot + plot_layout(heights = c(0.5, 3))
# dendrogram.vertical.plot <- plot_spacer() / module.dendrogram + plot_layout(heights = c(0.17, 1))
# combined.plot <- dendrogram.vertical.plot | vertical.plot
# combined.plot <- combined.plot + plot_layout(widths = c(0.2, 0.8)) + plot_annotation(
#   title = "Enrichment of Hallmark pathways across perturbations",
#   subtitle = paste(CELL.TYPE, GENE.AGGREGATION)
# )
# ggsave(paste(OUTPUT.DIR, "enrichment_bubble_plot_fgsea_2.pdf", sep = "/"),
#        combined.plot, height = 13, width = 13, device = pdf)
```

### Bubble plot, alphabetical - mostly for sgRNA_full

```{r}
combined.df <- bind_rows(fgsea.output.list, .id = "Perturbations")
combined.df$pathway <- gsub("HALLMARK_", "", combined.df$pathway)
combined.df$pathway <- gsub(" ", "", combined.df$pathway)
combined.df$log10pvalue <- -log10(combined.df$pval)
# combined.df$log10padj <- -log10(combined.df$padj)
combined.df$Perturbations <- gsub("_non-targeting_noRT", "", combined.df$Perturbations)
unfiltered.df <- combined.df

# combined.df <- combined.df %>%
#   group_by(pathway) %>%
#   filter(any(padj < 0.05)) %>%
#   ungroup()

combined.df <- combined.df %>%
  group_by(pathway) %>%
  filter(any(padj < 0.05)) %>%
  ungroup()

es.df <- combined.df %>%
  dplyr::select(-pval, -padj, -log2err, -NES, -size, -leadingEdge, -log10pvalue)

es.df <- es.df %>%
  group_by(pathway, Perturbations) %>%
  pivot_wider(names_from = Perturbations, values_from = ES) %>%
  group_by(pathway)

es.df <- as.data.frame(es.df)
rownames(es.df) <- es.df$pathway
es.df <- es.df %>% dplyr::select(-pathway)

df.ordered <- combined.df %>%
  arrange(match(pathway, rownames(es.df)), match(Perturbations, colnames(es.df)))

df.ordered$pathway <- factor(df.ordered$pathway, levels = rownames(es.df))
df.ordered$Perturbations <- factor(df.ordered$Perturbations, levels = colnames(es.df))

# Add significance layer
df.ordered$Significance <- ifelse(df.ordered$padj < 0.05, "adjp < 0.05", "NA")

# Clean up names
df.ordered$Pathways <- df.ordered$pathway
df.ordered$Pathways <- gsub("_", " ", df.ordered$Pathways)
```

Make the plot

```{r}
bubble.plot <- ggplot(df.ordered,
               aes(x = Perturbations)) +
  
  # Add the bubble and outline significance layers
  geom_point(aes(y = Pathways, size = log10pvalue, color = ES), alpha = 1.0) +
  geom_point(data = subset(df.ordered, padj < 0.05),
            aes(y = Pathways, size = log10pvalue, shape = 'adjp < 0.05'),
            color = "black", fill = NA, alpha = 0.8) +
  geom_point(data = subset(df.ordered, is.na(pval) | is.na(padj) | is.na(log10pvalue)), 
            aes(y = Pathways), 
            color = "black", size = 1, alpha = 0.6) +
  scale_size(range = c(2, 7)) +
  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "grey50"
  ) +
  scale_shape_manual(name = "Significance",
                   values = c(`adjp < 0.05` = 21),
                   guide = guide_legend(override.aes = list(color = "black", fill = NA))) +

  # Labels
  labs(
    color = "Effect size",
    size = "-Log10 padj"
  ) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.key.height = unit(0.3, 'cm'),
        axis.line = element_line(colour = "black"),
        legend.key = element_blank())

annotation.plot.theme <- theme(
  panel.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  axis.text.x = element_blank(),
  axis.title.x = element_blank(),
  axis.ticks.x = element_blank(),
  axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0))
)

combined.plot <- bubble.plot + plot_annotation(
  title = "Enrichment of Hallmark pathways across perturbations",
  subtitle = paste(CELL.TYPE, GENE.AGGREGATION)
)
ggsave(paste(OUTPUT.DIR, "enrichment_bubble_plot_fgsea_non-sorted.pdf", sep = "/"),
       combined.plot, height = 13, width = 13, device = pdf)
```

## Grouped cell type GSEA

Import all of the different cell types at the same time

```{r}
INPUT.DIR.BASE <- "/raleighlab/data1/czou/gbm_perturb/gbm_perturb_sb28_clean_outputs/deseq/SB28_integrated_20231207_ME/sgRNA_gene_sgNegCtrl3_knockouts"
INPUT.DIRS <- list.dirs(INPUT.DIR.BASE)[-1]
OUTPUT.DIR <- "/raleighlab/data1/czou/gbm_perturb/gbm_perturb_sb28_clean_outputs/gsea/sgRNA_gene_sgNegCtrl3_knockouts/Combined"
```

```{r}
data.list.symbols <- list()
for (dir in INPUT.DIRS) {
  file_names <- list.files(dir)
  split.string <- strsplit(dir, "/")[[1]]
  cell.type <- tail(split.string, 1)
  for (i in file_names) {
    name <- gsub(".csv", "", i)
    name <- paste(cell.type, name, sep = "_")
    df <- read.table(paste(dir, "/", i, sep = ""))
    data.list.symbols[[name]] <- df
  }
}
```

Prepare the ENSEMBL gene ID list, get rid of low expression, and order:

```{r}
# Prepare the ENSEMBL gene ID list
mapping.df <- annotables::grcm38
missing.genes <- list()
data.list.ensembl <- list()
for (name in names(data.list.symbols)) {
  deseq_table <- data.list.symbols[[name]]
  mapping.df <- mapping.df[!duplicated(mapping.df$symbol), ]
  mapping.table <- mapping.df[mapping.df$symbol %in% deseq_table$feature, ]
  mapping.table <- as.data.frame(mapping.table)
  rownames(mapping.table) <- mapping.table$symbol
  deseq_table$ensembl <- mapping.table[deseq_table$feature, "ensgene"]
  deseq_table.missing <- deseq_table[is.na(deseq_table$ensembl), ]
  missing.genes[[name]] <- deseq_table.missing$feature
  data.list.ensembl[[name]] <- deseq_table[!is.na(deseq_table$ensembl), ]
}
write.table(missing.genes[[1]], file = paste(OUTPUT.DIR, "missing_genes.txt", sep = "/"), row.names = FALSE, quote = FALSE)

# Get rid of low expression
processed.deseq.tables <- list()
for (name in names(data.list.ensembl)) {
  deseq_table <- data.list.ensembl[[name]]
  deseq_table <- deseq_table[!is.na(deseq_table$pvalue), ]
  deseq_table <- deseq_table[!is.na(deseq_table$padj), ]
  processed.deseq.tables[[name]] <- deseq_table
}
data.list <- processed.deseq.tables

# Create ordered lists
ranked_gene_set_list <- list()
for (name in names(processed.deseq.tables)) {
  deseq_table <- processed.deseq.tables[[name]]
  # Group by unique Ensembl ID and average
  deseq_table <- deseq_table %>%
    group_by(ensembl) %>%
    summarize(log_fc = mean(log_fc))
  deseq_table <- deseq_table %>% mutate(rank = rank(log_fc,  ties.method = "random"))
  deseq_table <- deseq_table[order(-deseq_table$rank),] # Rank deterministically
  
  # Generate a ranked list
  gene_list <- deseq_table$log_fc
  gene_names <- deseq_table$ensembl
  names(gene_list) <- gene_names
  ranked_gene_set_list[[name]] <- gene_list
}
```

Now run GSEA:

```{r}
msigdbr_df <- msigdbr(species = "Mus musculus", category = "H")
msigdbr_list = split(x = msigdbr_df$ensembl_gene, f = msigdbr_df$gs_name)

set.seed(1) # required for deterministic performance
fgsea.output.list <- list()
for (name in names(ranked_gene_set_list)) {
  gsea_results <- fgsea(pathways = msigdbr_list, stats = ranked_gene_set_list[[name]],
                        maxSize = 500, eps = 0.0) # allow arbitrarily low p-values
  fgsea.output.list[[name]] <- gsea_results
}
saveRDS(fgsea.output.list, paste(OUTPUT.DIR, "fgsea_ensembl_ontology_list.rds", sep = "/"))
```

### Heatmaps, clustered and alphabetical

```{r}
extracted_cols <- lapply(names(fgsea.output.list), function(name) {
  col <- fgsea.output.list[[name]]$ES
  names(col) <- fgsea.output.list[[name]]$pathway
  col
})
combined.df <- do.call(cbind, extracted_cols)
colnames(combined.df) <- names(fgsea.output.list)

ht = Heatmap(combined.df,
        name = "Enrichment Scores across Perturbations and Conditions",
        col = colorRamp2(c(-1, 0, 1), c("blue", "white", "red")),
        cluster_column_slices = FALSE,
        cluster_columns = FALSE,
        width = ncol(combined.df)*unit(4, "mm"),
        height = nrow(combined.df)*unit(4, "mm"),
)
pdf(paste(OUTPUT.DIR, "enrichment_combined_matrix_alphabetical.pdf", sep = "/"),
    width=35,
    height=19,
    useDingbats=FALSE)
draw(ht, heatmap_legend_side = "top", annotation_legend_side = "top")
dev.off()

extracted_cols <- lapply(names(fgsea.output.list), function(name) {
  col <- fgsea.output.list[[name]]$ES
  names(col) <- fgsea.output.list[[name]]$pathway
  col
})
combined.df <- do.call(cbind, extracted_cols)
colnames(combined.df) <- names(fgsea.output.list)

ht = Heatmap(combined.df,
        name = "Enrichment Scores across Perturbations and Conditions",
        col = colorRamp2(c(-1, 0, 1), c("blue", "white", "red")),
        cluster_column_slices = FALSE,
        cluster_columns = TRUE,
        width = ncol(combined.df)*unit(4, "mm"),
        height = nrow(combined.df)*unit(4, "mm"),
)
pdf(paste(OUTPUT.DIR, "enrichment_combined_matrix_clustered.pdf", sep = "/"),
    width=35,
    height=19,
    useDingbats=FALSE)
draw(ht, heatmap_legend_side = "top", annotation_legend_side = "top")
dev.off()
```

### Bubble plot, clustered across cell types

```{r}
# Eliminate Prkdc if needed
fgsea.output.list <- fgsea.output.list[-grep("Prkdc", names(fgsea.output.list))]

combined.df <- bind_rows(fgsea.output.list, .id = "Perturbations")
combined.df$pathway <- gsub("HALLMARK_", "", combined.df$pathway)
combined.df$pathway <- gsub(" ", "", combined.df$pathway)
combined.df$log10pvalue <- -log10(combined.df$padj)
combined.df$Perturbations <- gsub("_sgNegCtrl", "", combined.df$Perturbations)
unfiltered.df <- combined.df

# combined.df <- combined.df %>%
#   group_by(pathway) %>%
#   filter(any(padj < 0.05)) %>%
#   ungroup()

combined.df <- combined.df %>%
  group_by(pathway) %>%
  filter(any(padj < 0.05)) %>%
  ungroup()

es.df <- combined.df %>%
  dplyr::select(-pval, -padj, -log2err, -NES, -size, -leadingEdge, -log10pvalue)

es.df <- es.df %>%
  group_by(pathway, Perturbations) %>%
  pivot_wider(names_from = Perturbations, values_from = ES) %>%
  group_by(pathway)

es.df <- as.data.frame(es.df)
rownames(es.df) <- es.df$pathway
es.df <- es.df %>% dplyr::select(-pathway)

row_dist <- dist(es.df, method = "euclidean")
row_hclust <- hclust(row_dist, method = "complete")
row_order <- order.dendrogram(as.dendrogram(row_hclust))

col_dist <- dist(t(es.df), method = "euclidean")
col_hclust <- hclust(col_dist, method = "complete")
col_order <- order.dendrogram(as.dendrogram(col_hclust))

df.ordered <- combined.df %>%
  arrange(match(pathway, rownames(es.df)[row_order]), match(Perturbations, colnames(es.df)[col_order]))

df.ordered$pathway <- factor(df.ordered$pathway, levels = rownames(es.df)[row_order])
df.ordered$Perturbations <- factor(df.ordered$Perturbations, levels = colnames(es.df)[col_order])
unique_perturbations <- unique(df.ordered$Perturbations)

# Add in cell type fill bar

df.ordered$cell.type <- sapply(strsplit(as.character(df.ordered$Perturbations), "_"), `[`, 1)

rect.df <- data.frame(
  xmin = unique_perturbations,
  xmax = unique_perturbations,
  ymin = rep(max(as.numeric(df.ordered$pathway)) + 1, length(unique_perturbations)),
  ymax = rep(max(as.numeric(df.ordered$pathway)) + 2, length(unique_perturbations)),
  cell.type = df.ordered$cell.type[match(unique_perturbations, df.ordered$Perturbations)]
)
rect.df$numeric_xmin <- as.numeric(as.factor(rect.df$xmin)) - 0.5
rect.df$numeric_xmax <- as.numeric(as.factor(rect.df$xmax)) + 0.5

# Add in perturb fill bar

df.ordered$perturb <- sapply(strsplit(as.character(df.ordered$Perturbations), "_"), `[`, 2)

perturb.df <- data.frame(
  xmin = unique_perturbations,
  xmax = unique_perturbations,
  ymin = rep(max(as.numeric(df.ordered$pathway)) + 2, length(unique_perturbations)),
  ymax = rep(max(as.numeric(df.ordered$pathway)) + 3, length(unique_perturbations)),
  perturb = df.ordered$perturb[match(unique_perturbations, df.ordered$Perturbations)]
)
perturb.df$numeric_xmin <- as.numeric(as.factor(perturb.df$xmin)) - 0.5
perturb.df$numeric_xmax <- as.numeric(as.factor(perturb.df$xmax)) + 0.5

# Add significance layer
df.ordered$Significance <- ifelse(df.ordered$padj < 0.05, "adjp < 0.05", "NA")

# Clean up names
df.ordered$Pathways <- df.ordered$pathway
df.ordered$Pathways <- gsub("_", " ", df.ordered$Pathways)
```

```{r}
color.map <- list(
  `OPCs` = "#A6CEE3",
  `Oligodendrocytes` = "#1F78B4",
  `Microglia` = "#B2DF8A",
  `Macrophages` = "#33A02C",
  `Astrocytes` = "#FB9A99",
  `Lyz2` = "#E31A1C",
  `Apoe` = "#FDBF6F",
  `Cd44` = "#FF7F00",
  `C1qa` = "#CAB2D6",
  `Ptprc` = "#6A3D9A",
  `Cd74` = "#FFFF99",
  `Prkdc` = "brown"
)

bubble.plot <- ggplot(df.ordered,
               aes(x = Perturbations)) +
  
  # Add the bubble and outline significance layers
  geom_point(aes(y = Pathways, size = log10pvalue, color = ES), alpha = 1.0) +
  geom_point(data = subset(df.ordered, padj < 0.05),
            aes(y = Pathways, size = log10pvalue, shape = 'adjp < 0.05'),
            color = "black", fill = NA, alpha = 0.8) +
  geom_point(data = subset(df.ordered, is.na(pval) | is.na(padj) | is.na(log10pvalue)), 
            aes(y = Pathways), 
            color = "black", size = 1, alpha = 0.6) +
  scale_size(range = c(2, 7)) +
  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "grey50"
  ) +
  scale_shape_manual(name = "Significance",
                   values = c(`adjp < 0.05` = 21),
                   guide = guide_legend(override.aes = list(color = "black", fill = NA))) +
  
  # Add the cell type bar layer
  geom_rect(data = rect.df, aes(xmin = numeric_xmin, xmax = numeric_xmax, ymin = ymin, ymax = ymax, fill = cell.type), inherit.aes = FALSE) +
  
  # Add the perturb bar layer
  geom_rect(data = perturb.df, aes(xmin = numeric_xmin, xmax = numeric_xmax, ymin = ymin,
                                   ymax = ymax, fill = perturb), inherit.aes = FALSE) +
  
  # Color things in:
  scale_fill_manual(values = color.map) +

  # Labels
  labs(
    color = "Effect size",
    size = "-Log10 padj"
  ) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.key.height = unit(0.3, 'cm'),
        axis.line = element_line(colour = "black"),
        legend.key = element_blank())

perturb.dendrogram <- ggtree(as.phylo(col_hclust)) + theme_tree() + coord_flip() + scale_x_reverse()
module.dendrogram <- ggtree(as.phylo(row_hclust)) + theme_tree()

vertical.plot <- perturb.dendrogram / bubble.plot + plot_layout(heights = c(0.5, 3))
dendrogram.vertical.plot <- plot_spacer() / module.dendrogram + plot_layout(heights = c(0.22, 1))
combined.plot <- dendrogram.vertical.plot | vertical.plot
combined.plot <- combined.plot + plot_layout(widths = c(0.2, 0.8)) + plot_annotation(
  title = "Enrichment of Hallmark pathways clustered across cell types",
  subtitle = "Prkdc excluded"
)
ggsave(paste(OUTPUT.DIR, "enrichment_bubble_plot_fgsea_clustered_prkdc_excluded_2.pdf", sep = "/"),
       combined.plot, height = 9, width = 10, device = pdf)
```

### Bubble plot, but clustered within cell types

```{r}
# Eliminate Prkdc if required
fgsea.output.list <- fgsea.output.list[-grep("Prkdc", names(fgsea.output.list))]

combined.df <- bind_rows(fgsea.output.list, .id = "Perturbations")
combined.df$pathway <- gsub("HALLMARK_", "", combined.df$pathway)
combined.df$pathway <- gsub(" ", "", combined.df$pathway)
combined.df$log10pvalue <- -log10(combined.df$padj)
combined.df$Perturbations <- gsub("_sgNegCtrl", "", combined.df$Perturbations)
unfiltered.df <- combined.df

# combined.df <- combined.df %>%
#   group_by(pathway) %>%
#   filter(any(padj < 0.05)) %>%
#   ungroup()

combined.df <- combined.df %>%
  group_by(pathway) %>%
  filter(any(padj < 0.05)) %>%
  ungroup()

es.df <- combined.df %>%
  dplyr::select(-pval, -padj, -log2err, -NES, -size, -leadingEdge, -log10pvalue)

es.df <- es.df %>%
  group_by(pathway, Perturbations) %>%
  pivot_wider(names_from = Perturbations, values_from = ES) %>%
  group_by(pathway)

es.df <- as.data.frame(es.df)
rownames(es.df) <- es.df$pathway
es.df <- es.df %>% dplyr::select(-pathway)

row_dist <- dist(es.df, method = "euclidean")
row_hclust <- hclust(row_dist, method = "complete")
row_order <- order.dendrogram(as.dendrogram(row_hclust))

# Group the cells within the cell types
ordering <- c()
for (cell.type in c("Astrocytes", "Macrophages", "Microglia", "OPCs", "Oligodendrocytes")) {
  cols <- grep(sprintf("^%s_", cell.type), colnames(es.df), value = TRUE)
  col.df <- es.df[cols]
  col.dist <- dist(t(col.df), method = "euclidean")
  col.hclust <- hclust(col.dist, method = "complete")
  col.order <- order.dendrogram(as.dendrogram(col.hclust))
  ordering <- c(ordering, cols[col.order])
}
es.df <- es.df[, ordering]

df.ordered <- combined.df %>%
  arrange(match(pathway, rownames(es.df)[row_order]), match(Perturbations, colnames(es.df)))

df.ordered$pathway <- factor(df.ordered$pathway, levels = rownames(es.df)[row_order])
df.ordered$Perturbations <- factor(df.ordered$Perturbations, levels = colnames(es.df))
unique_perturbations <- unique(df.ordered$Perturbations)

# Add in cell type fill bar

df.ordered$cell.type <- sapply(strsplit(as.character(df.ordered$Perturbations), "_"), `[`, 1)

rect.df <- data.frame(
  xmin = unique_perturbations,
  xmax = unique_perturbations,
  ymin = rep(max(as.numeric(df.ordered$pathway)) + 1, length(unique_perturbations)),
  ymax = rep(max(as.numeric(df.ordered$pathway)) + 2, length(unique_perturbations)),
  cell.type = df.ordered$cell.type[match(unique_perturbations, df.ordered$Perturbations)]
)
rect.df$numeric_xmin <- as.numeric(as.factor(rect.df$xmin)) - 0.5
rect.df$numeric_xmax <- as.numeric(as.factor(rect.df$xmax)) + 0.5

# Add in perturb fill bar

df.ordered$perturb <- sapply(strsplit(as.character(df.ordered$Perturbations), "_"), `[`, 2)

perturb.df <- data.frame(
  xmin = unique_perturbations,
  xmax = unique_perturbations,
  ymin = rep(max(as.numeric(df.ordered$pathway)) + 2, length(unique_perturbations)),
  ymax = rep(max(as.numeric(df.ordered$pathway)) + 3, length(unique_perturbations)),
  perturb = df.ordered$perturb[match(unique_perturbations, df.ordered$Perturbations)]
)
perturb.df$numeric_xmin <- as.numeric(as.factor(perturb.df$xmin)) - 0.5
perturb.df$numeric_xmax <- as.numeric(as.factor(perturb.df$xmax)) + 0.5

# Add significance layer
df.ordered$Significance <- ifelse(df.ordered$padj < 0.05, "adjp < 0.05", "NA")

# Clean up names
df.ordered$Pathways <- df.ordered$pathway
df.ordered$Pathways <- gsub("_", " ", df.ordered$Pathways)
```

```{r}
color.map <- list(
  `OPCs` = "#A6CEE3",
  `Oligodendrocytes` = "#1F78B4",
  `Microglia` = "#B2DF8A",
  `Macrophages` = "#33A02C",
  `Astrocytes` = "#FB9A99",
  `Lyz2` = "#E31A1C",
  `Apoe` = "#FDBF6F",
  `Cd44` = "#FF7F00",
  `C1qa` = "#CAB2D6",
  `Ptprc` = "#6A3D9A",
  `Cd74` = "#FFFF99",
  `Prkdc` = "brown"
)
bubble.plot <- ggplot(df.ordered,
               aes(x = Perturbations)) +
  
  # Add the bubble and outline significance layers
  geom_point(aes(y = Pathways, size = log10pvalue, color = ES), alpha = 1.0) +
  geom_point(data = subset(df.ordered, padj < 0.05),
            aes(y = Pathways, size = log10pvalue, shape = 'adjp < 0.05'),
            color = "black", fill = NA, alpha = 0.8) +
  geom_point(data = subset(df.ordered, is.na(pval) | is.na(padj) | is.na(log10pvalue)), 
            aes(y = Pathways), 
            color = "black", size = 1, alpha = 0.6) +
  scale_size(range = c(2, 7)) +
  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "grey50"
  ) +
  scale_shape_manual(name = "Significance",
                   values = c(`adjp < 0.05` = 21),
                   guide = guide_legend(override.aes = list(color = "black", fill = NA))) +
  
  # Add the cell type bar layer
  geom_rect(data = rect.df, aes(xmin = numeric_xmin, xmax = numeric_xmax, ymin = ymin, ymax = ymax, fill = cell.type), inherit.aes = FALSE) +
  
  # Add the perturb bar layer
  geom_rect(data = perturb.df, aes(xmin = numeric_xmin, xmax = numeric_xmax, ymin = ymin,
                                   ymax = ymax, fill = perturb), inherit.aes = FALSE) +
  
  # Color things in:
  scale_fill_manual(values = color.map) +

  # Labels
  labs(
    color = "Effect size",
    size = "-Log10 padj"
  ) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.key.height = unit(0.3, 'cm'),
        axis.line = element_line(colour = "black"),
        legend.key = element_blank())

module.dendrogram <- ggtree(as.phylo(row_hclust)) + theme_tree()

vertical.plot <- bubble.plot
dendrogram.vertical.plot <- plot_spacer() / module.dendrogram + plot_layout(heights = c(0.01, 1))
combined.plot <- dendrogram.vertical.plot | vertical.plot
combined.plot <- combined.plot + plot_layout(widths = c(0.2, 0.8)) + plot_annotation(
  title = "Enrichment of Hallmark pathways clustered within cell types",
  subtitle = "Prkdc excluded"
)
ggsave(paste(OUTPUT.DIR, "enrichment_bubble_plot_fgsea_clustered_within_cells_prkdc_excluded_2.pdf", sep = "/"),
       combined.plot, height = 10, width = 9, device = pdf)
```

## Heatmap of top-20 genes by LFC

Import all of the data

```{r}
INPUT.DIR.BASE <- "/raleighlab/data1/czou/gbm_perturb/gbm_perturb_sb28_clean_outputs/deseq/SB28_integrated_20231207_ME/sgRNA_gene_sgNegCtrl3_knockouts"
INPUT.DIRS <- list.dirs(INPUT.DIR.BASE)[-1]
OUTPUT.DIR <- "/raleighlab/data1/czou/gbm_perturb/gbm_perturb_sb28_clean_outputs/gsea/sgRNA_gene_sgNegCtrl3_knockouts/Combined"
```

```{r}
data.list.cell.types <- list()
for (dir in INPUT.DIRS) {
  file_names <- list.files(dir)
  split.string <- strsplit(dir, "/")[[1]]
  cell.type <- tail(split.string, 1)
  data.list.symbols <- list()
  for (i in file_names) {
    name <- gsub("_sgNegCtrl.csv", "", i)
    df <- read.table(paste(dir, "/", i, sep = ""))
    data.list.symbols[[name]] <- df
  }
  data.list.cell.types[[cell.type]] <- data.list.symbols
}
data.list.cell.types[["SB28"]] <- NULL
```

For consistency's sake, filter the dataframes by the ENSEMBL and DESeq criteria we identified before with GSEA.

```{r}
# Prepare the ENSEMBL gene ID list
mapping.df <- annotables::grcm38
missing.genes <- list()
for (cell.type in names(data.list.cell.types)) {
  data.list.cell.type <- data.list.cell.types[[cell.type]]
  for (name in names(data.list.cell.type)) {
    deseq_table <- data.list.cell.type[[name]]
    mapping.df <- mapping.df[!duplicated(mapping.df$symbol), ]
    mapping.table <- as.data.frame(mapping.table)
    rownames(mapping.table) <- mapping.table$symbol
    deseq_table.missing <- deseq_table[is.na(deseq_table$ensembl), ]
    missing.genes[[name]] <- deseq_table.missing$feature
    data.list.cell.type[[name]] <- deseq_table[!is.na(deseq_table$ensembl), ]
    
    deseq_table <- deseq_table[!is.na(deseq_table$pvalue), ]
    deseq_table <- deseq_table[!is.na(deseq_table$padj), ]
    data.list.cell.type[[name]] <- deseq_table    
  }
  data.list.cell.types[[cell.type]] <- data.list.cell.type
}
```

For each perturbation, construct a dataframe that consists of the top 20 upregulated genes by LFC with LFC \> 0.5. Then mash the dataframes together and plot them on a heatmap.

```{r}
# Get the top genes per perturbation
perturbations <- unique(unlist(data.list.cell.types))
top_genes_per_perturbation <- list()

for (perturbation in perturbations) {
  all_top_genes <- vector("list", length(data.list.cell.types))

  for (i in seq_along(data.list.cell.types)) {
    cell_type_list <- data.list.cell.types[[i]]
    df <- cell_type_list[[perturbation]]
    df <- df[df$log_fc > 0.5, ]
    sorted_genes <- df[order(df$log_fc, decreasing = TRUE), "feature"]
    top_20_genes <- head(sorted_genes, 20)
    all_top_genes[[i]] <- top_20_genes
  }
  unique_genes <- unique(unlist(all_top_genes))
  top_genes_per_perturbation[[perturbation]] <- unique_genes
}

# For each perturbation, construct a new dataframe containing
# those top genes across cell type x perturbation

dataframes_per_perturbation <- list()

# Create column names: combination of each cell type and each perturbation
all_cell_types <- names(data.list.cell.types)
all_combinations <- expand.grid(all_cell_types, perturbations)
col_names <- apply(all_combinations, 1, paste, collapse = "_")

# Iterate over each perturbation
for (perturbation in perturbations) {
  top_genes <- top_genes_per_perturbation[[perturbation]]
  rows <- length(top_genes)

  # Create an empty dataframe with the appropriate column names
  df <- data.frame(matrix(nrow = rows, ncol = length(col_names), dimnames = list(top_genes, col_names)))
  
  # Populate the dataframe
  for (cell_type in all_cell_types) {
    cell_type_list <- data.list.cell.types[[cell_type]]
    for (perturb in perturbations) {
      col_name <- paste(cell_type, perturb, sep = "_")
      for (gene in top_genes) {
        deseq_df <- cell_type_list[[perturb]]
        log_fc <- deseq_df[deseq_df$feature == gene, "log_fc"]
        if (length(log_fc) == 0) {
          log_fc <- NA
        }
        df[gene, col_name] <- log_fc
      }
    }
  }
  dataframes_per_perturbation[[perturbation]] <- df
}
```

Now, merge all of the dataframes together by column name and make a heatmap.

```{r}
merged_df <- NULL
for (perturbation in names(dataframes_per_perturbation)) {
  df <- dataframes_per_perturbation[[perturbation]]
  rownames(df) <- paste(rownames(df), perturbation, sep = "_")
  merged_df <- if (is.null(merged_df)) df else rbind(merged_df, df)
}
```

```{r}
split_col_names <- strsplit(colnames(merged_df), "_")
cell_types <- sapply(split_col_names, `[[`, 1)
perturbations <- sapply(split_col_names, `[[`, 2)

cell.type.color.scheme <- c(
  `OPCs` = "#A6CEE3",
  `Oligodendrocytes` = "#1F78B4",
  `Microglia` = "#B2DF8A",
  `Macrophages` = "#33A02C",
  `Astrocytes` = "#FB9A99"
)

perturbation.color.scheme <- c(
  `Prkdc` = "#E31A1C",
  `Apoe` = "#FDBF6F",
  `C1qa` = "#FF7F00",
  `Cd44` = "#CAB2D6",
  `Cd74` = "#6A3D9A",
  `Lyz2` = "#FFFF99",
  `Ptprc` = "#B15928"
)

# Create a bar on the top for perturbation x cell type
col_annotation <- HeatmapAnnotation(
  df = data.frame(Cell_Type = cell_types, Perturbation = perturbations),
  col = list(Cell_Type = cell.type.color.scheme, 
             Perturbation = perturbation.color.scheme),
  which = "col")

# Create a bar on the right side for source perturbation
source_perturbation <- sapply(strsplit(rownames(merged_df), "_"), tail, 1)
row_annotation <- HeatmapAnnotation(source = source_perturbation,
                                    col = list(source = perturbation.color.scheme),
                                    which = "row")

# Plot the heatmap
pdf(paste(OUTPUT.DIR, "top20_perturbation_matrix.pdf", sep = "/"),
    width=20,
    height=25,
    useDingbats=FALSE)
matrix <- as.matrix(merged_df)
rownames(matrix) <- NULL
ht <- Heatmap(matrix, 
        top_annotation = col_annotation, 
        right_annotation = row_annotation,
        name = "Top-20 log FC across perturbations and cell types",
        cluster_column_slices = FALSE,
        cluster_columns = FALSE,
        cluster_rows = FALSE,
        na_col = "gray",
        width = ncol(merged_df)*unit(4, "mm"),
        height = nrow(merged_df)*unit(0.5, "mm"),
)
draw(ht, heatmap_legend_side = "top", annotation_legend_side = "top")
dev.off()
```
