# R-GSEA - GL261 Edition

Set libpaths

```{r}
# Set libpaths
.libPaths(c("/raleighlab/data1/czou/gbm_perturb/gbm_perturb/renv/library/R-4.3/x86_64-pc-linux-gnu", .libPaths()))
```

Import libraries

```{r}
library(Seurat)
library(clusterProfiler)
library(org.Hs.eg.db)
library(knitr)
library(msigdbr)
library(fgsea)
library(dplyr)
library(data.table)
library(ggplot2)
library(grid)
library(ComplexHeatmap)
library(colorRamp2)
library(tidyr)
library(patchwork)
library(ggtree)
library(gridExtra)
library(ape)
library(RColorBrewer)
library(reactome.db)
library(annotables)
library(biomaRt)
library(scales)
set.seed(5220)
```

Define a bunch of constants.

```{r}
INPUT_DIRS <- c("/raleighlab/data1/czou/gbm_perturb/gbm_perturb_gl261_clean_outputs/deseq/condNormalized_CED")
CONTEXT <- "CED"
CONDITION <- "Combined"
OUTPUT_DIR <- sprintf("/raleighlab/data1/czou/gbm_perturb/gbm_perturb_gl261_clean_outputs/gsea/condNormalized/%s/%s", CONTEXT, CONDITION)
```

Import the DESeq output

```{r}
data.list <- list()
for (dir in INPUT_DIRS) {
  file_names <- list.files(dir)
  for (i in file_names) {
    name <- gsub(".csv", "", i)
    df <- read.table(paste(dir, "/", i, sep = ""))
    data.list[[name]] <- df
  }
}
```

Separate out RT and noRT use cases

```{r}
noRT.indices <- grepl("_noRT_non-targeting_noRT", names(data.list))
RT.indices <- grepl("_RT_non-targeting_RT", names(data.list))
noRT.list <- data.list[noRT.indices]
RT.list <- data.list[RT.indices]
overall.list <- data.list

if (CONDITION == "RT") {
  data.list.symbols <- RT.list
} else if (CONDITION == "Combined") {
  data.list.symbols <- overall.list
} else {
  data.list.symbols <- noRT.list
}
```

Prepare an ENSEMBL gene ID data list by converting using the `annotables` package. Record the genes that end up going missing from ENSEMBL conversion.

```{r}
mapping.df <- annotables::grcm38
missing.genes <- list()
data.list.ensembl <- list()
for (name in names(data.list.symbols)) {
  deseq_table <- data.list.symbols[[name]]
  mapping.df <- mapping.df[!duplicated(mapping.df$symbol), ]
  mapping.table <- mapping.df[mapping.df$symbol %in% deseq_table$feature, ]
  mapping.table <- as.data.frame(mapping.table)
  rownames(mapping.table) <- mapping.table$symbol
  deseq_table$ensembl <- mapping.table[deseq_table$feature, "ensgene"]
  deseq_table.missing <- deseq_table[is.na(deseq_table$ensembl), ]
  missing.genes[[name]] <- deseq_table.missing$feature
  data.list.ensembl[[name]] <- deseq_table[!is.na(deseq_table$ensembl), ]
}
write.table(missing.genes[[1]], file = paste(OUTPUT_DIR, "missing_genes.txt", sep = "/"), row.names = FALSE, quote = FALSE)
```

Get rid of anything that shows low expression:

```{r}
processed.deseq.tables <- list()
for (name in names(data.list.ensembl)) {
  deseq_table <- data.list.ensembl[[name]]
  deseq_table <- deseq_table[!is.na(deseq_table$pvalue), ]
  deseq_table <- deseq_table[!is.na(deseq_table$padj), ]
  processed.deseq.tables[[name]] <- deseq_table
}
data.list <- processed.deseq.tables
```

Generate an ordered list from the processed DESeq2 tables:

```{r}
ranked_gene_set_list <- list()
for (name in names(processed.deseq.tables)) {
  deseq_table <- processed.deseq.tables[[name]]
  # Group by unique Ensembl ID and average
  deseq_table <- deseq_table %>%
    group_by(ensembl) %>%
    summarize(log_fc = mean(log_fc))
  deseq_table <- deseq_table %>% mutate(rank = rank(log_fc,  ties.method = "random"))
  deseq_table <- deseq_table[order(-deseq_table$rank),] # Rank deterministically
  
  # Generate a ranked list
  gene_list <- deseq_table$log_fc
  gene_names <- deseq_table$ensembl
  names(gene_list) <- gene_names
  ranked_gene_set_list[[name]] <- gene_list
}
```

Run GSEA using `msigdbr` and `fgsea`.

```{r}
msigdbr_df <- msigdbr(species = "Mus musculus", category = "H")
msigdbr_list = split(x = msigdbr_df$ensembl_gene, f = msigdbr_df$gs_name)

fgsea.output.list <- list()
for (name in names(ranked_gene_set_list)) {
  gsea_results <- fgsea(pathways = msigdbr_list, stats = ranked_gene_set_list[[name]],
                        maxSize = 500, eps = 0.0) # allow arbitrarily low p-values
  fgsea.output.list[[name]] <- gsea_results
}
saveRDS(fgsea.output.list, paste(OUTPUT_DIR, "fgsea_ensembl_ontology_list_nes.rds", sep = "/"))
```

Plot the enrichment results on a heatmap

```{r}
extracted_cols <- lapply(names(fgsea.output.list), function(name) {
  col <- fgsea.output.list[[name]]$NES
  names(col) <- fgsea.output.list[[name]]$pathway
  col
})
combined.df <- do.call(cbind, extracted_cols)
colnames(combined.df) <- names(fgsea.output.list)

ht = Heatmap(combined.df,
        name = "Enrichment Scores across Perturbations and Conditions",
        col = colorRamp2(c(-3, 0, 3), c("blue", "white", "red")),
        cluster_column_slices = FALSE,
        cluster_columns = TRUE,
        width = ncol(combined.df)*unit(4, "mm"),
        height = nrow(combined.df)*unit(4, "mm"),
)
pdf(paste(OUTPUT_DIR, "enrichment_combined_matrix_nes.pdf", sep = "/"),
    width=35,
    height=19,
    useDingbats=FALSE)
draw(ht, heatmap_legend_side = "top", annotation_legend_side = "top")
dev.off()
```

### Bubble Plot, grouped across radiation

```{r}
combined.df <- bind_rows(fgsea.output.list, .id = "Perturbations")
combined.df$pathway <- gsub("HALLMARK_", "", combined.df$pathway)
combined.df$pathway <- gsub(" ", "", combined.df$pathway)
combined.df$log10pvalue <- -log10(combined.df$padj)
combined.df$Perturbations <- gsub("_non-targeting_noRT", "", combined.df$Perturbations)
combined.df$Perturbations <- gsub("_non-targeting_RT", "", combined.df$Perturbations)
unfiltered.df <- combined.df

combined.df <- combined.df %>%
  group_by(pathway) %>%
  filter(any(padj < 0.05)) %>%
  ungroup()

es.df <- combined.df %>%
  dplyr::select(-pval, -padj, -log2err, -ES, -size, -leadingEdge, -log10pvalue)

es.df <- es.df %>%
  group_by(pathway, Perturbations) %>%
  pivot_wider(names_from = Perturbations, values_from = NES) %>%
  group_by(pathway)

es.df <- as.data.frame(es.df)
rownames(es.df) <- es.df$pathway
es.df <- es.df %>% dplyr::select(-pathway)

row_dist <- dist(es.df, method = "euclidean")
row_hclust <- hclust(row_dist, method = "complete")
row_order <- order.dendrogram(as.dendrogram(row_hclust))

col_dist <- dist(t(es.df), method = "euclidean")
col_hclust <- hclust(col_dist, method = "complete")
col_order <- order.dendrogram(as.dendrogram(col_hclust))

df.ordered <- combined.df %>%
  arrange(match(pathway, rownames(es.df)[row_order]), match(Perturbations, colnames(es.df)[col_order]))

df.ordered$pathway <- factor(df.ordered$pathway, levels = rownames(es.df)[row_order])
df.ordered$Perturbations <- factor(df.ordered$Perturbations, levels = colnames(es.df)[col_order])

# Add in RT fill bar

df.ordered$RT <- ifelse(grepl("_RT", df.ordered$Perturbations), "RT", "noRT")
unique_perturbations <- unique(df.ordered$Perturbations)

rt.df <- data.frame(
  xmin = unique_perturbations,
  xmax = unique_perturbations,
  ymin = rep(max(as.numeric(df.ordered$pathway)) + 1, length(unique_perturbations)),
  ymax = rep(max(as.numeric(df.ordered$pathway)) + 2, length(unique_perturbations)),
  RT = ifelse(df.ordered$RT[match(unique_perturbations, df.ordered$Perturbations)] == "RT", "RT", "noRT")
)
rt.df$numeric_xmin <- as.numeric(as.factor(rt.df$xmin)) - 0.5
rt.df$numeric_xmax <- as.numeric(as.factor(rt.df$xmax)) + 0.5

unique_perturbations <- unique(gsub("_noRT", "", unique_perturbations))
unique_perturbations <- unique(gsub("_RT", "", unique_perturbations))

# Add in GO fill bar

go_categories_RT = read.table("/raleighlab/data1/liuj/gbm_perturb/analysis/features_48h_GL261_annot2.csv", sep = ",")
colnames(go_categories_RT) = c("Gene", "GO")
rownames(go_categories_RT) = go_categories_RT$Gene
rownames(go_categories_RT) <- paste(CONTEXT, rownames(go_categories_RT), "RT", sep = "_")
rownames(go_categories_RT) <- gsub("NTC", "non-targeting", rownames(go_categories_RT))

go_categories_noRT = read.table("/raleighlab/data1/liuj/gbm_perturb/analysis/features_48h_GL261_annot2.csv", sep = ",")
colnames(go_categories_noRT) = c("Gene", "GO")
rownames(go_categories_noRT) = go_categories_noRT$Gene
rownames(go_categories_noRT) <- paste(CONTEXT, rownames(go_categories_noRT), "noRT", sep = "_")
rownames(go_categories_noRT) <- gsub("NTC", "non-targeting", rownames(go_categories_noRT))

go_categories <- rbind(go_categories_RT, go_categories_noRT)
df.ordered$GO <- go_categories[as.character(df.ordered$Perturbations), "GO"]
unique_perturbations <- unique(df.ordered$Perturbations)

rect.df <- data.frame(
  xmin = unique_perturbations,
  xmax = unique_perturbations,
  ymin = rep(max(as.numeric(df.ordered$pathway)) + 2, length(unique_perturbations)),
  ymax = rep(max(as.numeric(df.ordered$pathway)) + 3, length(unique_perturbations)),
  GO = go_categories[as.character(unique_perturbations), "GO"]
)
rect.df$numeric_xmin <- as.numeric(as.factor(rect.df$xmin)) - 0.5
rect.df$numeric_xmax <- as.numeric(as.factor(rect.df$xmax)) + 0.5

# Hard coded color mapping for consistency

term.color.map <- list(
  `Mitosis` = "#A6CEE3",
  `Translation` = "#1F78B4",
  `DNA damage response` = "#B2DF8A",
  `Metabolism` = "#33A02C",
  `LIM Domain` = "#FB9A99",
  `Mitochondria` = "#E31A1C",
  `Proteasome` = "#FDBF6F",
  `Signaling` = "#FF7F00",
  `Cyclin-dependent kinase inhibitor` = "#CAB2D6",
  `DNA Replication` = "#6A3D9A",
  `Telomere` = "#FFFF99",
  `Transcription` = "#B15928",
  `Non-targeting` = "gray",
  `RT` = "darkred",
  `noRT` = 'darkblue'
)

# Add in screening phenotype annotations

phenoTbl <- read.table("/raleighlab/data1/liuj/gbm_perturb/analysis/GL261_1+2_results_table.txt",sep='\t',header=TRUE,row.names=1)
coverageTbl <- read.table('/raleighlab/data1/liuj/gbm_perturb/analysis/gbm_pdx_perturb_GL261_integrate_xfp/pdx_perturb_GL261_concordant_sgRNAs.txt',header=TRUE,sep='\t')
coverageTblCmp <- rbind(invitronoRT = apply(coverageTbl[c("GL261_48hit_noRT_1","GL261_48hit_noRT_2"),],2,sum),
                        invitroRT = apply(coverageTbl[c("GL261_48hit_RT_1","GL261_48hit_RT_2"),],2,sum),
                        preinfnoRT = apply(coverageTbl[c("GL261_noRT_preinf_MACSFACS_1","GL261_noRT_preinf_MACSFACS_2","GL261_noRT_preinf_MACSFACS_3"),],2,sum),
                        preinfRT = apply(coverageTbl[c("GL261_RT_preinf_MACSFACS_1","GL261_RT_preinf_MACSFACS_2","GL261_RT_preinf_MACSFACS_3"),],2,sum),
                        CEDnoRT = apply(coverageTbl[c("GL261_CED_pool","GL261_noRT_CED_MACSFACS"),],2,sum),
                        CEDRT = coverageTbl[c("GL261_RT_CED_MACSFACS"),])

colnames(coverageTblCmp) <- gsub("non.targeting","NTC",colnames(coverageTblCmp))
sgAnnot <- data.frame(row.names = colnames(coverageTblCmp), gamma = phenoTbl[colnames(coverageTblCmp),"gamma"], tau = phenoTbl[colnames(coverageTblCmp),"tau"], rho = phenoTbl[colnames(coverageTblCmp),"rho"])
sgAnnot["non-targeting",] <- c(0,0,0)
sgAnnot <- sgAnnot[!(rownames(sgAnnot) %in% c("NTC", "NTC_B")),]

RT.sgAnnot <- sgAnnot
rownames(RT.sgAnnot) <- paste(rownames(RT.sgAnnot), "_RT", sep = "")

noRT.sgAnnot <- sgAnnot
rownames(noRT.sgAnnot) <- paste(rownames(noRT.sgAnnot), "_noRT", sep = "")

annotation.table <- rbind(RT.sgAnnot, noRT.sgAnnot)
annotation.table <- annotation.table[rownames(annotation.table) != "non-targeting_noRT", ]

rownames(annotation.table) <- paste(CONTEXT, rownames(annotation.table), sep = "_")
annotation.table <- subset(annotation.table, 
                           rownames(annotation.table) %in% unique(df.ordered$Perturbations))
annotation.table$Perturbations <- rownames(annotation.table)
annotation.table$Perturbations <- factor(annotation.table$Perturbations, 
                                         levels = levels(df.ordered$Perturbations))
annotation.table <- annotation.table %>%
  arrange(Perturbations)

# Add significance layer
df.ordered$Significance <- ifelse(df.ordered$padj < 0.05, "adjp < 0.05", "NA")

# Clean up names
df.ordered$Pathways <- df.ordered$pathway
df.ordered$Pathways <- gsub("_", " ", df.ordered$Pathways)
```

Then make the plot:

```{r}
bubble.plot <- ggplot(df.ordered,
               aes(x = Perturbations)) +
  
  # Add the bubble and outline significance layers
  geom_point(aes(y = Pathways, size = log10pvalue, color = NES), alpha = 1.0) +
  geom_point(data = subset(df.ordered, padj < 0.05),
            aes(y = Pathways, size = log10pvalue, shape = 'adjp < 0.05'),
            color = "black", fill = NA, alpha = 0.8) +
  geom_point(data = subset(df.ordered, is.na(pval) | is.na(padj) | is.na(log10pvalue)), 
            aes(y = Pathways), 
            color = "black", size = 1, alpha = 0.6) +
  scale_size(range = c(2, 7)) +
  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "grey50"
  ) +
  scale_shape_manual(name = "Significance",
                   values = c(`adjp < 0.05` = 21),
                   guide = guide_legend(override.aes = list(color = "black", fill = NA))) +
  
  # Add the GO layer
  geom_rect(data = rect.df, aes(xmin = numeric_xmin, xmax = numeric_xmax, ymin = ymin, ymax = ymax, fill = GO), inherit.aes = FALSE) +
  scale_fill_manual(values = term.color.map) +
  
  # Add the RT layer
  geom_rect(data = rt.df, aes(xmin = numeric_xmin, xmax = numeric_xmax, ymin = ymin, ymax = ymax, fill = RT), inherit.aes = FALSE) +

  # Labels
  labs(
    color = "Effect size",
    size = "-Log10 padj"
  ) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.key.height = unit(0.3, 'cm'),
        axis.line = element_line(colour = "black"),
        legend.key = element_blank())

annotation.plot.theme <- theme(
  panel.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  axis.text.x = element_blank(),
  axis.title.x = element_blank(),
  axis.ticks.x = element_blank(),
  axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0))
)
gamma.plot <- ggplot(annotation.table, aes(x = Perturbations, y = gamma)) +
  geom_col(width = 1.0, fill = "grey") + annotation.plot.theme
tau.plot <- ggplot(annotation.table, aes(x = Perturbations, y = tau)) +
  geom_col(width = 1.0, fill = "grey") + annotation.plot.theme
rho.plot <- ggplot(annotation.table, aes(x = Perturbations, y = rho)) +
  geom_col(width = 1.0, fill = "grey") + annotation.plot.theme

perturb.dendrogram <- ggtree(as.phylo(col_hclust)) + theme_tree() + coord_flip() + scale_x_reverse()
module.dendrogram <- ggtree(as.phylo(row_hclust)) + theme_tree()

vertical.plot <- gamma.plot / tau.plot / rho.plot / perturb.dendrogram / bubble.plot + plot_layout(heights = c(0.5, 0.5, 0.5, 0.5, 3))
dendrogram.vertical.plot <- plot_spacer() / module.dendrogram + plot_layout(heights = c(0.85, 1))
combined.plot <- dendrogram.vertical.plot | vertical.plot
combined.plot <- combined.plot + plot_layout(widths = c(0.2, 0.8)) + plot_annotation(
  title = "Enrichment of Hallmark pathways across perturbations",
  subtitle = paste(CONTEXT, CONDITION, "condNormalized")
)
ggsave(paste(OUTPUT_DIR, "enrichment_bubble_plot_fgsea_nes_2.pdf", sep = "/"),
       combined.plot, height = 13, 
       width = 6.7 + (length(unique(df.ordered$Perturbations)) / 27 * (11 - 6.7)) / 0.8, device = pdf)
```

### Bubble plot, grouped within radiation conditions:

```{r}
combined.df <- bind_rows(fgsea.output.list, .id = "Perturbations")
combined.df$pathway <- gsub("HALLMARK_", "", combined.df$pathway)
combined.df$pathway <- gsub(" ", "", combined.df$pathway)
combined.df$log10pvalue <- -log10(combined.df$padj)
combined.df$Perturbations <- gsub("_non-targeting_noRT", "", combined.df$Perturbations)
combined.df$Perturbations <- gsub("_non-targeting_RT", "", combined.df$Perturbations)
unfiltered.df <- combined.df

combined.df <- combined.df %>%
  group_by(pathway) %>%
  filter(any(padj < 0.05)) %>%
  ungroup()

es.df <- combined.df %>%
  dplyr::select(-pval, -padj, -log2err, -ES, -size, -leadingEdge, -log10pvalue)

es.df <- es.df %>%
  group_by(pathway, Perturbations) %>%
  pivot_wider(names_from = Perturbations, values_from = NES) %>%
  group_by(pathway)

es.df <- as.data.frame(es.df)
rownames(es.df) <- es.df$pathway
es.df <- es.df %>% dplyr::select(-pathway)

row_dist <- dist(es.df, method = "euclidean")
row_hclust <- hclust(row_dist, method = "complete")
row_order <- order.dendrogram(as.dendrogram(row_hclust))

ordering <- c()
col.hclusts <- list()
for (rt.condition in c("RT", "noRT")) {
  cols <- grep(sprintf("_%s$", rt.condition), colnames(es.df), value = TRUE)
  col.df <- es.df[cols]
  col.dist <- dist(t(col.df), method = "euclidean")
  col.hclust <- hclust(col.dist, method = "complete")
  col.hclusts[[rt.condition]] <- col.hclust
  col.order <- order.dendrogram(as.dendrogram(col.hclust))
  ordering <- c(ordering, cols[col.order])
}
es.df <- es.df[, ordering]

df.ordered <- combined.df %>%
  arrange(match(pathway, rownames(es.df)[row_order]), match(Perturbations, colnames(es.df)))

df.ordered$pathway <- factor(df.ordered$pathway, levels = rownames(es.df)[row_order])
df.ordered$Perturbations <- factor(df.ordered$Perturbations, levels = colnames(es.df))

# Add in RT fill bar

df.ordered$RT <- ifelse(grepl("_RT", df.ordered$Perturbations), "RT", "noRT")
unique_perturbations <- unique(df.ordered$Perturbations)

rt.df <- data.frame(
  xmin = unique_perturbations,
  xmax = unique_perturbations,
  ymin = rep(max(as.numeric(df.ordered$pathway)) + 1, length(unique_perturbations)),
  ymax = rep(max(as.numeric(df.ordered$pathway)) + 2, length(unique_perturbations)),
  RT = ifelse(df.ordered$RT[match(unique_perturbations, df.ordered$Perturbations)] == "RT", "RT", "noRT")
)
rt.df$numeric_xmin <- as.numeric(as.factor(rt.df$xmin)) - 0.5
rt.df$numeric_xmax <- as.numeric(as.factor(rt.df$xmax)) + 0.5

unique_perturbations <- unique(gsub("_noRT", "", unique_perturbations))
unique_perturbations <- unique(gsub("_RT", "", unique_perturbations))

# Add in GO fill bar

go_categories_RT = read.table("/raleighlab/data1/liuj/gbm_perturb/analysis/features_48h_GL261_annot2.csv", sep = ",")
colnames(go_categories_RT) = c("Gene", "GO")
rownames(go_categories_RT) = go_categories_RT$Gene
rownames(go_categories_RT) <- paste(CONTEXT, rownames(go_categories_RT), "RT", sep = "_")
rownames(go_categories_RT) <- gsub("NTC", "non-targeting", rownames(go_categories_RT))

go_categories_noRT = read.table("/raleighlab/data1/liuj/gbm_perturb/analysis/features_48h_GL261_annot2.csv", sep = ",")
colnames(go_categories_noRT) = c("Gene", "GO")
rownames(go_categories_noRT) = go_categories_noRT$Gene
rownames(go_categories_noRT) <- paste(CONTEXT, rownames(go_categories_noRT), "noRT", sep = "_")
rownames(go_categories_noRT) <- gsub("NTC", "non-targeting", rownames(go_categories_noRT))

go_categories <- rbind(go_categories_RT, go_categories_noRT)
df.ordered$GO <- go_categories[as.character(df.ordered$Perturbations), "GO"]
unique_perturbations <- unique(df.ordered$Perturbations)

rect.df <- data.frame(
  xmin = unique_perturbations,
  xmax = unique_perturbations,
  ymin = rep(max(as.numeric(df.ordered$pathway)) + 2, length(unique_perturbations)),
  ymax = rep(max(as.numeric(df.ordered$pathway)) + 3, length(unique_perturbations)),
  GO = go_categories[as.character(unique_perturbations), "GO"]
)
rect.df$numeric_xmin <- as.numeric(as.factor(rect.df$xmin)) - 0.5
rect.df$numeric_xmax <- as.numeric(as.factor(rect.df$xmax)) + 0.5

# Hard coded color mapping for consistency

term.color.map <- list(
  `Mitosis` = "#A6CEE3",
  `Translation` = "#1F78B4",
  `DNA damage response` = "#B2DF8A",
  `Metabolism` = "#33A02C",
  `LIM Domain` = "#FB9A99",
  `Mitochondria` = "#E31A1C",
  `Proteasome` = "#FDBF6F",
  `Signaling` = "#FF7F00",
  `Cyclin-dependent kinase inhibitor` = "#CAB2D6",
  `DNA Replication` = "#6A3D9A",
  `Telomere` = "#FFFF99",
  `Transcription` = "#B15928",
  `Non-targeting` = "gray",
  `RT` = "darkred",
  `noRT` = 'darkblue'
)

# Add in screening phenotype annotations

phenoTbl <- read.table("/raleighlab/data1/liuj/gbm_perturb/analysis/GL261_1+2_results_table.txt",sep='\t',header=TRUE,row.names=1)
coverageTbl <- read.table('/raleighlab/data1/liuj/gbm_perturb/analysis/gbm_pdx_perturb_GL261_integrate_xfp/pdx_perturb_GL261_concordant_sgRNAs.txt',header=TRUE,sep='\t')
coverageTblCmp <- rbind(invitronoRT = apply(coverageTbl[c("GL261_48hit_noRT_1","GL261_48hit_noRT_2"),],2,sum),
                        invitroRT = apply(coverageTbl[c("GL261_48hit_RT_1","GL261_48hit_RT_2"),],2,sum),
                        preinfnoRT = apply(coverageTbl[c("GL261_noRT_preinf_MACSFACS_1","GL261_noRT_preinf_MACSFACS_2","GL261_noRT_preinf_MACSFACS_3"),],2,sum),
                        preinfRT = apply(coverageTbl[c("GL261_RT_preinf_MACSFACS_1","GL261_RT_preinf_MACSFACS_2","GL261_RT_preinf_MACSFACS_3"),],2,sum),
                        CEDnoRT = apply(coverageTbl[c("GL261_CED_pool","GL261_noRT_CED_MACSFACS"),],2,sum),
                        CEDRT = coverageTbl[c("GL261_RT_CED_MACSFACS"),])

colnames(coverageTblCmp) <- gsub("non.targeting","NTC",colnames(coverageTblCmp))
sgAnnot <- data.frame(row.names = colnames(coverageTblCmp), gamma = phenoTbl[colnames(coverageTblCmp),"gamma"], tau = phenoTbl[colnames(coverageTblCmp),"tau"], rho = phenoTbl[colnames(coverageTblCmp),"rho"])
sgAnnot["non-targeting",] <- c(0,0,0)
sgAnnot <- sgAnnot[!(rownames(sgAnnot) %in% c("NTC", "NTC_B")),]

RT.sgAnnot <- sgAnnot
rownames(RT.sgAnnot) <- paste(rownames(RT.sgAnnot), "_RT", sep = "")

noRT.sgAnnot <- sgAnnot
rownames(noRT.sgAnnot) <- paste(rownames(noRT.sgAnnot), "_noRT", sep = "")

annotation.table <- rbind(RT.sgAnnot, noRT.sgAnnot)
annotation.table <- annotation.table[rownames(annotation.table) != "non-targeting_noRT", ]

rownames(annotation.table) <- paste(CONTEXT, rownames(annotation.table), sep = "_")
annotation.table <- subset(annotation.table, 
                           rownames(annotation.table) %in% unique(df.ordered$Perturbations))
annotation.table$Perturbations <- rownames(annotation.table)
annotation.table$Perturbations <- factor(annotation.table$Perturbations, 
                                         levels = levels(df.ordered$Perturbations))
annotation.table <- annotation.table %>%
  arrange(Perturbations)

# Add significance layer
df.ordered$Significance <- ifelse(df.ordered$padj < 0.05, "adjp < 0.05", "NA")

# Clean up names
df.ordered$Pathways <- df.ordered$pathway
levels(df.ordered$Pathways) <- gsub("_", " ", levels(df.ordered$Pathways))

new_levels <- levels(df.ordered$Perturbations)
new_levels <- gsub("CED_", "", new_levels)
new_levels <- gsub("preinf_", "", new_levels)
new_levels <- gsub("invitro_", "", new_levels)
levels(df.ordered$Perturbations) <- new_levels
```

Make the plot

```{r}
bubble.plot <- ggplot(df.ordered,
               aes(x = Perturbations)) +
  
  # Add the bubble and outline significance layers
  geom_point(aes(y = Pathways, size = log10pvalue, color = NES), alpha = 1.0) +
  geom_point(data = subset(df.ordered, padj < 0.05),
            aes(y = Pathways, size = log10pvalue, shape = 'adjp < 0.05'),
            color = "black", fill = NA, alpha = 0.8) +
  geom_point(data = subset(df.ordered, is.na(pval) | is.na(padj) | is.na(log10pvalue)), 
            aes(y = Pathways), 
            color = "black", size = 1, alpha = 0.6) +
  scale_size(range = c(2, 7)) +
  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "grey50"
  ) +
  scale_shape_manual(name = "Significance",
                   values = c(`adjp < 0.05` = 21),
                   guide = guide_legend(override.aes = list(color = "black", fill = NA))) +
  
  # Add the GO layer
  geom_rect(data = rect.df, aes(xmin = numeric_xmin, xmax = numeric_xmax, ymin = ymin, ymax = ymax, fill = GO), inherit.aes = FALSE) +
  scale_fill_manual(values = term.color.map) +
  
  # Add the RT layer
  geom_rect(data = rt.df, aes(xmin = numeric_xmin, xmax = numeric_xmax, ymin = ymin, ymax = ymax, fill = RT), inherit.aes = FALSE) +

  # Labels
  labs(
    color = "Effect size",
    size = "-Log10 padj"
  ) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.key.height = unit(0.3, 'cm'),
        axis.line = element_line(colour = "black"),
        legend.key = element_blank())

annotation.plot.theme <- theme(
  panel.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  axis.text.x = element_blank(),
  axis.title.x = element_blank(),
  axis.ticks.x = element_blank(),
  axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0))
)
gamma.plot <- ggplot(annotation.table, aes(x = Perturbations, y = gamma)) +
  geom_col(width = 1.0, fill = "grey") + annotation.plot.theme
tau.plot <- ggplot(annotation.table, aes(x = Perturbations, y = tau)) +
  geom_col(width = 1.0, fill = "grey") + annotation.plot.theme
rho.plot <- ggplot(annotation.table, aes(x = Perturbations, y = rho)) +
  geom_col(width = 1.0, fill = "grey") + annotation.plot.theme

perturb.dendrogram.RT <- ggtree(as.phylo(col.hclusts[["RT"]])) + theme_tree() + coord_flip() + scale_x_reverse()
perturb.dendrogram.noRT <- ggtree(as.phylo(col.hclusts[["noRT"]])) + theme_tree() + coord_flip() + scale_x_reverse()
module.dendrogram <- ggtree(as.phylo(row_hclust)) + theme_tree()

vertical.plot <- gamma.plot / tau.plot / rho.plot / (perturb.dendrogram.RT | perturb.dendrogram.noRT) / bubble.plot + plot_layout(heights = c(0.5, 0.5, 0.5, 0.5, 3))
dendrogram.vertical.plot <- plot_spacer() / module.dendrogram + plot_layout(heights = c(0.85, 1))
combined.plot <- dendrogram.vertical.plot | vertical.plot
combined.plot <- combined.plot + plot_layout(widths = c(0.2, 0.8)) + plot_annotation(
  title = "Enrichment of Hallmark pathways across perturbations",
  subtitle = paste(CONTEXT, CONDITION, "condNormalized")
)
ggsave(paste(OUTPUT_DIR, "enrichment_bubble_plot_fgsea_nes_within_RT_all_clustered.pdf", sep = "/"),
       combined.plot, height = 13, 
       width = 6.7 + (length(unique(df.ordered$Perturbations)) / 27 * (11 - 6.7)) / 0.8, device = pdf)
```

#### Overlay

Overlay LFCs over the existing plot. This means that we need to add an additional column to the dataframe. In this column we will have the average LFC for all genes in the intersection between:

1.  That pathway's genes

2.  The union of DE genes across all perturbations and conditions, where a gene is differentially expressed if it has abs(LFC) \> 0.1 and adjusted p-value less than 0.05.

```{r}
if (CONDITION != "Combined") {
  print("Can't run this section if not combined!")
}
processed_deseq_tables <- list()
for (name in names(data.list.ensembl)) {
  deseq_table <- data.list.ensembl[[name]]
  deseq_table <- deseq_table[!is.na(deseq_table$pvalue), ]
  deseq_table <- deseq_table[!is.na(deseq_table$padj), ]
  deseq_table <- deseq_table[deseq_table$padj < 0.05, ]
  deseq_table <- deseq_table[abs(deseq_table$log_fc) > 0.1, ]
  processed_deseq_tables[[name]] <- deseq_table
}
data_list <- processed_deseq_tables

union_de_genes <- c()
for (name in names(data_list)) {
  deseq_table <- data_list[[name]]
  ensembl_ids <- deseq_table$ensembl
  union_de_genes <- unique(c(union_de_genes, ensembl_ids))
}
```

Note that this list isn't exactly equal to the DE gene plots at the moment because we also filter out all genes that don't have a valid ENSEMBL ID. Next, we get the list of DESeq dataframes and filter them for these genes.

```{r}
filter_to_de_genes <- function(deseq_table) {
  deseq_table <- subset(deseq_table, ensembl %in% union_de_genes)
}
de_gene_data_list <- lapply(data.list.ensembl, filter_to_de_genes)
```

Generate an `msigdbr_df` that represents the genes representing each pathway.

```{r}
msigdbr_df <- msigdbr(species = "Mus musculus", category = "H")
msigdbr_list = split(x = msigdbr_df$ensembl_gene, f = msigdbr_df$gs_name)
```

For each perturbation x condition in the `de_gene_data_list`, build a DESeq dataframe that includes only the intersection between the pathway's genes and the current ones.

```{r}
perturb_condition_pathway_dfs <- list()
for (perturb_condition in names(de_gene_data_list)) {
  deseq_df <- de_gene_data_list[[perturb_condition]]
  for (pathway in names(msigdbr_list)) {
    pathway_intersected_df <- subset(deseq_df, ensembl %in% msigdbr_list[[pathway]])
    perturb_condition_pathway <- paste(perturb_condition, pathway, sep = "_")
    perturb_condition_pathway_dfs[[perturb_condition_pathway]] <- pathway_intersected_df
  }
}
```

Find the average LFC across dataframe in `perturb_condition_pathway_dfs`. Note that for any gene that doesn't pass expression filters, we set its contribution to the average to 0.

```{r}
calculate_average <- function(deseq_table) {
  deseq_table$log_fc[is.na(deseq_table$pvalue) & is.na(deseq_table$padj)] <- 0
  return(mean(deseq_table$log_fc))
}
perturb_condition_pathway_means <- lapply(perturb_condition_pathway_dfs, calculate_average)
```

Map the averages to the `df.ordered` table generated above

```{r}
# Generate a mapping column for the df.ordered table
df.ordered$perturb_cond_pathway <- paste(df.ordered$Perturbations, df.ordered$Pathways, sep = "_")
df.ordered$perturb_cond_pathway <- gsub(" ", "_", df.ordered$perturb_cond_pathway)

# Generate the corresponding mapping for the means
names(perturb_condition_pathway_means) <- gsub("non-targeting_RT_HALLMARK_|non-targeting_noRT_HALLMARK_|invitro_|CED_|preinf_", "", names(perturb_condition_pathway_means))

df.ordered$perturb_cond_pathway_means <- unlist(perturb_condition_pathway_means[df.ordered$perturb_cond_pathway])
```

Make the plot, but instead of using NES as the value, use the mapped means.

```{r}
bubble.plot <- ggplot(df.ordered,
               aes(x = Perturbations)) +
  
  # Add the bubble and outline significance layers
  geom_point(aes(y = Pathways, size = log10pvalue, color = perturb_cond_pathway_means), alpha = 1.0) +
  geom_point(data = subset(df.ordered, padj < 0.05),
            aes(y = Pathways, size = log10pvalue, shape = 'adjp < 0.05'),
            color = "black", fill = NA, alpha = 0.8) +
  geom_point(data = subset(df.ordered, is.na(pval) | is.na(padj) | is.na(log10pvalue)), 
            aes(y = Pathways), 
            color = "black", size = 1, alpha = 0.6) +
  scale_size(range = c(2, 7)) +
  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "grey50",
    limits = c(-0.75, 0.75),
    oob = scales::squish
  ) +
  scale_shape_manual(name = "Significance",
                   values = c(`adjp < 0.05` = 21),
                   guide = guide_legend(override.aes = list(color = "black", fill = NA))) +
  
  # Add the GO layer
  geom_rect(data = rect.df, aes(xmin = numeric_xmin, xmax = numeric_xmax, ymin = ymin, ymax = ymax, fill = GO), inherit.aes = FALSE) +
  scale_fill_manual(values = term.color.map) +
  
  # Add the RT layer
  geom_rect(data = rt.df, aes(xmin = numeric_xmin, xmax = numeric_xmax, ymin = ymin, ymax = ymax, fill = RT), inherit.aes = FALSE) +

  # Labels
  labs(
    color = "Effect size",
    size = "-Log10 padj"
  ) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.key.height = unit(0.3, 'cm'),
        axis.line = element_line(colour = "black"),
        legend.key = element_blank())

annotation.plot.theme <- theme(
  panel.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  axis.text.x = element_blank(),
  axis.title.x = element_blank(),
  axis.ticks.x = element_blank(),
  axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0))
)
gamma.plot <- ggplot(annotation.table, aes(x = Perturbations, y = gamma)) +
  geom_col(width = 1.0, fill = "grey") + annotation.plot.theme
tau.plot <- ggplot(annotation.table, aes(x = Perturbations, y = tau)) +
  geom_col(width = 1.0, fill = "grey") + annotation.plot.theme
rho.plot <- ggplot(annotation.table, aes(x = Perturbations, y = rho)) +
  geom_col(width = 1.0, fill = "grey") + annotation.plot.theme

perturb.dendrogram.RT <- ggtree(as.phylo(col.hclusts[["RT"]])) + theme_tree() + coord_flip() + scale_x_reverse()
perturb.dendrogram.noRT <- ggtree(as.phylo(col.hclusts[["noRT"]])) + theme_tree() + coord_flip() + scale_x_reverse()
module.dendrogram <- ggtree(as.phylo(row_hclust)) + theme_tree()

vertical.plot <- gamma.plot / tau.plot / rho.plot / (perturb.dendrogram.RT | perturb.dendrogram.noRT) / bubble.plot + plot_layout(heights = c(0.5, 0.5, 0.5, 0.5, 3))
dendrogram.vertical.plot <- plot_spacer() / module.dendrogram + plot_layout(heights = c(0.85, 1))
combined.plot <- dendrogram.vertical.plot | vertical.plot
combined.plot <- combined.plot + plot_layout(widths = c(0.2, 0.8)) + plot_annotation(
  title = "Enrichment of Hallmark pathways across perturbations",
  subtitle = paste(CONTEXT, CONDITION, "condNormalized")
)
ggsave(paste(OUTPUT_DIR, "enrichment_bubble_plot_fgsea_nes_within_RT_logfc_overlay.pdf", sep = "/"),
       combined.plot, height = 13, 
       width = 6.7 + (length(unique(df.ordered$Perturbations)) / 27 * (11 - 6.7)) / 0.8, device = pdf)
```

For each pathway, figure out which genes we want

For each perturbation/condition for each pathway, build a dataframe that includes only the intersection between the pathway's genes and the filtered DESeq output

Find the average LFC across each pathway/perturbation_condition dataframe,

### Bubble plot, but alphabetical

Create an additional version where the perturbations are ordered alphabetically:

```{r}
combined.df <- bind_rows(fgsea.output.list, .id = "Perturbations")
combined.df$pathway <- gsub("HALLMARK_", "", combined.df$pathway)
combined.df$pathway <- gsub(" ", "", combined.df$pathway)
combined.df$log10pvalue <- -log10(combined.df$padj)
combined.df$Perturbations <- gsub("_non-targeting_noRT", "", combined.df$Perturbations)
combined.df$Perturbations <- gsub("_non-targeting_RT", "", combined.df$Perturbations)
unfiltered.df <- combined.df

combined.df <- combined.df %>%
  group_by(pathway) %>%
  filter(any(padj < 0.05)) %>%
  ungroup()

es.df <- combined.df %>%
  dplyr::select(-pval, -padj, -log2err, -ES, -size, -leadingEdge, -log10pvalue)

es.df <- es.df %>%
  group_by(pathway, Perturbations) %>%
  pivot_wider(names_from = Perturbations, values_from = NES) %>%
  group_by(pathway)

es.df <- as.data.frame(es.df)
rownames(es.df) <- es.df$pathway
es.df <- es.df %>% dplyr::select(-pathway)

row_dist <- dist(es.df, method = "euclidean")
row_hclust <- hclust(row_dist, method = "complete")
row_order <- order.dendrogram(as.dendrogram(row_hclust))

df.ordered <- combined.df %>%
  arrange(match(pathway, rownames(es.df)[row_order]), match(Perturbations, colnames(es.df)))

df.ordered$pathway <- factor(df.ordered$pathway, levels = rownames(es.df)[row_order])
df.ordered$Perturbations <- factor(df.ordered$Perturbations, levels = sort(colnames(es.df)))

# Add in RT fill bar

df.ordered$RT <- ifelse(grepl("_RT", df.ordered$Perturbations), "RT", "noRT")
unique_perturbations <- unique(df.ordered$Perturbations)

rt.df <- data.frame(
  xmin = unique_perturbations,
  xmax = unique_perturbations,
  ymin = rep(max(as.numeric(df.ordered$pathway)) + 1, length(unique_perturbations)),
  ymax = rep(max(as.numeric(df.ordered$pathway)) + 2, length(unique_perturbations)),
  RT = ifelse(df.ordered$RT[match(unique_perturbations, df.ordered$Perturbations)] == "RT", "RT", "noRT")
)
rt.df$numeric_xmin <- as.numeric(as.factor(rt.df$xmin)) - 0.5
rt.df$numeric_xmax <- as.numeric(as.factor(rt.df$xmax)) + 0.5

unique_perturbations <- unique(gsub("_noRT", "", unique_perturbations))
unique_perturbations <- unique(gsub("_RT", "", unique_perturbations))

# Add in GO fill bar

go_categories_RT = read.table("/raleighlab/data1/liuj/gbm_perturb/analysis/features_48h_GL261_annot2.csv", sep = ",")
colnames(go_categories_RT) = c("Gene", "GO")
rownames(go_categories_RT) = go_categories_RT$Gene
rownames(go_categories_RT) <- paste(CONTEXT, rownames(go_categories_RT), "RT", sep = "_")
rownames(go_categories_RT) <- gsub("NTC", "non-targeting", rownames(go_categories_RT))

go_categories_noRT = read.table("/raleighlab/data1/liuj/gbm_perturb/analysis/features_48h_GL261_annot2.csv", sep = ",")
colnames(go_categories_noRT) = c("Gene", "GO")
rownames(go_categories_noRT) = go_categories_noRT$Gene
rownames(go_categories_noRT) <- paste(CONTEXT, rownames(go_categories_noRT), "noRT", sep = "_")
rownames(go_categories_noRT) <- gsub("NTC", "non-targeting", rownames(go_categories_noRT))

go_categories <- rbind(go_categories_RT, go_categories_noRT)
df.ordered$GO <- go_categories[as.character(df.ordered$Perturbations), "GO"]
unique_perturbations <- unique(df.ordered$Perturbations)

rect.df <- data.frame(
  xmin = unique_perturbations,
  xmax = unique_perturbations,
  ymin = rep(max(as.numeric(df.ordered$pathway)) + 2, length(unique_perturbations)),
  ymax = rep(max(as.numeric(df.ordered$pathway)) + 3, length(unique_perturbations)),
  GO = go_categories[as.character(unique_perturbations), "GO"]
)
rect.df$numeric_xmin <- as.numeric(as.factor(rect.df$xmin)) - 0.5
rect.df$numeric_xmax <- as.numeric(as.factor(rect.df$xmax)) + 0.5

# Hard coded color mapping for consistency

term.color.map <- list(
  `Mitosis` = "#A6CEE3",
  `Translation` = "#1F78B4",
  `DNA damage response` = "#B2DF8A",
  `Metabolism` = "#33A02C",
  `LIM Domain` = "#FB9A99",
  `Mitochondria` = "#E31A1C",
  `Proteasome` = "#FDBF6F",
  `Signaling` = "#FF7F00",
  `Cyclin-dependent kinase inhibitor` = "#CAB2D6",
  `DNA Replication` = "#6A3D9A",
  `Telomere` = "#FFFF99",
  `Transcription` = "#B15928",
  `Non-targeting` = "gray",
  `RT` = "darkred",
  `noRT` = 'darkblue'
)

# Add in screening phenotype annotations

phenoTbl <- read.table("/raleighlab/data1/liuj/gbm_perturb/analysis/GL261_1+2_results_table.txt",sep='\t',header=TRUE,row.names=1)
coverageTbl <- read.table('/raleighlab/data1/liuj/gbm_perturb/analysis/gbm_pdx_perturb_GL261_integrate_xfp/pdx_perturb_GL261_concordant_sgRNAs.txt',header=TRUE,sep='\t')
coverageTblCmp <- rbind(invitronoRT = apply(coverageTbl[c("GL261_48hit_noRT_1","GL261_48hit_noRT_2"),],2,sum),
                        invitroRT = apply(coverageTbl[c("GL261_48hit_RT_1","GL261_48hit_RT_2"),],2,sum),
                        preinfnoRT = apply(coverageTbl[c("GL261_noRT_preinf_MACSFACS_1","GL261_noRT_preinf_MACSFACS_2","GL261_noRT_preinf_MACSFACS_3"),],2,sum),
                        preinfRT = apply(coverageTbl[c("GL261_RT_preinf_MACSFACS_1","GL261_RT_preinf_MACSFACS_2","GL261_RT_preinf_MACSFACS_3"),],2,sum),
                        CEDnoRT = apply(coverageTbl[c("GL261_CED_pool","GL261_noRT_CED_MACSFACS"),],2,sum),
                        CEDRT = coverageTbl[c("GL261_RT_CED_MACSFACS"),])

colnames(coverageTblCmp) <- gsub("non.targeting","NTC",colnames(coverageTblCmp))
sgAnnot <- data.frame(row.names = colnames(coverageTblCmp), gamma = phenoTbl[colnames(coverageTblCmp),"gamma"], tau = phenoTbl[colnames(coverageTblCmp),"tau"], rho = phenoTbl[colnames(coverageTblCmp),"rho"])
sgAnnot["non-targeting",] <- c(0,0,0)
sgAnnot <- sgAnnot[!(rownames(sgAnnot) %in% c("NTC", "NTC_B")),]

RT.sgAnnot <- sgAnnot
rownames(RT.sgAnnot) <- paste(rownames(RT.sgAnnot), "_RT", sep = "")

noRT.sgAnnot <- sgAnnot
rownames(noRT.sgAnnot) <- paste(rownames(noRT.sgAnnot), "_noRT", sep = "")

annotation.table <- rbind(RT.sgAnnot, noRT.sgAnnot)
annotation.table <- annotation.table[rownames(annotation.table) != "non-targeting_noRT", ]

rownames(annotation.table) <- paste(CONTEXT, rownames(annotation.table), sep = "_")
annotation.table <- subset(annotation.table, 
                           rownames(annotation.table) %in% unique(df.ordered$Perturbations))
annotation.table$Perturbations <- rownames(annotation.table)
annotation.table$Perturbations <- factor(annotation.table$Perturbations, 
                                         levels = levels(df.ordered$Perturbations))
annotation.table <- annotation.table %>%
  arrange(Perturbations)

# Add significance layer
df.ordered$Significance <- ifelse(df.ordered$padj < 0.05, "adjp < 0.05", "NA")

# Clean up names
df.ordered$Pathways <- df.ordered$pathway
df.ordered$Pathways <- gsub("_", " ", df.ordered$Pathways)
# df.ordered$Perturbations <- gsub("CED_", "", df.ordered$Perturbations)
# df.ordered$Perturbations <- gsub("preinf_", "", df.ordered$Perturbations)
# df.ordered$Perturbations <- gsub("invitro_", "", df.ordered$Perturbations)
```

```{r}
bubble.plot <- ggplot(df.ordered,
               aes(x = Perturbations)) +
  
  # Add the bubble and outline significance layers
  geom_point(aes(y = Pathways, size = log10pvalue, color = NES), alpha = 1.0) +
  geom_point(data = subset(df.ordered, padj < 0.05),
            aes(y = Pathways, size = log10pvalue, shape = 'adjp < 0.05'),
            color = "black", fill = NA, alpha = 0.8) +
  geom_point(data = subset(df.ordered, is.na(pval) | is.na(padj) | is.na(log10pvalue)), 
            aes(y = Pathways), 
            color = "black", size = 1, alpha = 0.6) +
  scale_size(range = c(2, 7)) +
  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "grey50"
  ) +
  scale_shape_manual(name = "Significance",
                   values = c(`adjp < 0.05` = 21),
                   guide = guide_legend(override.aes = list(color = "black", fill = NA))) +
  
  # Add the GO layer
  geom_rect(data = rect.df, aes(xmin = numeric_xmin, xmax = numeric_xmax, ymin = ymin, ymax = ymax, fill = GO), inherit.aes = FALSE) +
  scale_fill_manual(values = term.color.map) +
  
  # Add the RT layer
  geom_rect(data = rt.df, aes(xmin = numeric_xmin, xmax = numeric_xmax, ymin = ymin, ymax = ymax, fill = RT), inherit.aes = FALSE) +

  # Labels
  labs(
    color = "Effect size",
    size = "-Log10 padj"
  ) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.key.height = unit(0.3, 'cm'),
        axis.line = element_line(colour = "black"),
        legend.key = element_blank())

annotation.plot.theme <- theme(
  panel.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  axis.text.x = element_blank(),
  axis.title.x = element_blank(),
  axis.ticks.x = element_blank(),
  axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0))
)
gamma.plot <- ggplot(annotation.table, aes(x = Perturbations, y = gamma)) +
  geom_col(width = 1.0, fill = "grey") + annotation.plot.theme
tau.plot <- ggplot(annotation.table, aes(x = Perturbations, y = tau)) +
  geom_col(width = 1.0, fill = "grey") + annotation.plot.theme
rho.plot <- ggplot(annotation.table, aes(x = Perturbations, y = rho)) +
  geom_col(width = 1.0, fill = "grey") + annotation.plot.theme

perturb.dendrogram.RT <- ggtree(as.phylo(col.hclusts[["RT"]])) + theme_tree() + coord_flip() + scale_x_reverse()
perturb.dendrogram.noRT <- ggtree(as.phylo(col.hclusts[["noRT"]])) + theme_tree() + coord_flip() + scale_x_reverse()
module.dendrogram <- ggtree(as.phylo(row_hclust)) + theme_tree()

vertical.plot <- gamma.plot / tau.plot / rho.plot / bubble.plot + plot_layout(heights = c(0.5, 0.5, 0.5, 3))
dendrogram.vertical.plot <- plot_spacer() / module.dendrogram + plot_layout(heights = c(0.85, 1))
combined.plot <- dendrogram.vertical.plot | vertical.plot
combined.plot <- combined.plot + plot_layout(widths = c(0.2, 0.8)) + plot_annotation(
  title = "Enrichment of Hallmark pathways across perturbations",
  subtitle = paste(CONTEXT, CONDITION, "condNormalized")
)
ggsave(paste(OUTPUT_DIR, "enrichment_bubble_plot_fgsea_nes_alphabetical.pdf", sep = "/"),
       combined.plot, height = 13, 
       width = 6.7 + (length(unique(df.ordered$Perturbations)) / 27 * (11 - 6.7)) / 0.8, device = pdf)
```

Keeping track of what dimensions we want for each combination:

```{r}
# CED noRT, size 2-6
# vertical.plot <- gamma.plot / tau.plot / rho.plot / perturb.dendrogram / bubble.plot + plot_layout(heights = c(0.5, 0.5, 0.5, 0.5, 3))
# dendrogram.vertical.plot <- plot_spacer() / module.dendrogram + plot_layout(heights = c(0.85, 1))
# combined.plot <- dendrogram.vertical.plot | vertical.plot
# combined.plot <- combined.plot + plot_layout(widths = c(0.2, 0.8)) + plot_annotation(
#   title = "Enrichment of Hallmark pathways across perturbations",
#   subtitle = paste(CONTEXT, CONDITION, "noRTNormalized")
# )
# ggsave(paste(OUTPUT_DIR, "enrichment_bubble_plot_fgsea.pdf", sep = "/"),
#        combined.plot, height = 13, width = 14, device = pdf)

# CED RT, size 2-7
# vertical.plot <- gamma.plot / tau.plot / rho.plot / perturb.dendrogram / bubble.plot + plot_layout(heights = c(0.5, 0.5, 0.5, 0.5, 3))
# dendrogram.vertical.plot <- plot_spacer() / module.dendrogram + plot_layout(heights = c(0.85, 1))
# combined.plot <- dendrogram.vertical.plot | vertical.plot
# combined.plot <- combined.plot + plot_layout(widths = c(0.2, 0.8)) + plot_annotation(
#   title = "Enrichment of Hallmark pathways across perturbations",
#   subtitle = paste(CONTEXT, CONDITION, "noRTNormalized")
# )
# ggsave(paste(OUTPUT_DIR, "enrichment_bubble_plot_fgsea.pdf", sep = "/"),
#        combined.plot, height = 13, width = 12, device = pdf)

# preinf RT, size 2-7
# vertical.plot <- gamma.plot / tau.plot / rho.plot / perturb.dendrogram / bubble.plot + plot_layout(heights = c(0.5, 0.5, 0.5, 0.5, 3))
# dendrogram.vertical.plot <- plot_spacer() / module.dendrogram + plot_layout(heights = c(0.85, 1))
# combined.plot <- dendrogram.vertical.plot | vertical.plot
# combined.plot <- combined.plot + plot_layout(widths = c(0.2, 0.8)) + plot_annotation(
#   title = "Enrichment of Hallmark pathways across perturbations",
#   subtitle = paste(CONTEXT, CONDITION, "noRTNormalized")
# )
# ggsave(paste(OUTPUT_DIR, "enrichment_bubble_plot_fgsea.pdf", sep = "/"),
#        combined.plot, height = 13, width = 13, device = pdf)
# preinf noRT same as above

# invitro noRT
# vertical.plot <- gamma.plot / tau.plot / rho.plot / perturb.dendrogram / bubble.plot + plot_layout(heights = c(0.5, 0.5, 0.5, 0.5, 3))
# dendrogram.vertical.plot <- plot_spacer() / module.dendrogram + plot_layout(heights = c(0.85, 1))
# combined.plot <- dendrogram.vertical.plot | vertical.plot
# combined.plot <- combined.plot + plot_layout(widths = c(0.2, 0.8)) + plot_annotation(
#   title = "Enrichment of Hallmark pathways across perturbations",
#   subtitle = paste(CONTEXT, CONDITION, "noRTNormalized")
# )
# ggsave(paste(OUTPUT_DIR, "enrichment_bubble_plot_fgsea.pdf", sep = "/"),
#        combined.plot, height = 13, width = 14, device = pdf)

# invitro RT
# vertical.plot <- gamma.plot / tau.plot / rho.plot / perturb.dendrogram / bubble.plot + plot_layout(heights = c(0.5, 0.5, 0.5, 0.5, 3))
# dendrogram.vertical.plot <- plot_spacer() / module.dendrogram + plot_layout(heights = c(0.85, 1))
# combined.plot <- dendrogram.vertical.plot | vertical.plot
# combined.plot <- combined.plot + plot_layout(widths = c(0.2, 0.8)) + plot_annotation(
#   title = "Enrichment of Hallmark pathways across perturbations",
#   subtitle = paste(CONTEXT, CONDITION, "noRTNormalized")
# )
# ggsave(paste(OUTPUT_DIR, "enrichment_bubble_plot_fgsea.pdf", sep = "/"),
#        combined.plot, height = 13, width = 14, device = pdf)
```

### Basic Analysis Questions:

```{r}
analysis.df <- merge(unfiltered.df, annotation.table, on = "Perturbations")
analysis.df$leadingEdge <- sapply(analysis.df$leadingEdge, function(x) paste(x, collapse = ","))
write.table(analysis.df, file = paste(OUTPUT_DIR, "analysis_df.txt", sep = "/"), row.names = FALSE, quote = FALSE)

# For each perturbation, how many significant hallmark gene sets are there?
perturb.grouped <- analysis.df %>%
  group_by(Perturbations) %>%
  summarize(n_hallmark_significant = sum(padj < 0.05, na.rm = TRUE)) %>%
  complete(Perturbations, fill = list(n_hallmark_significant = 0))
write.table(perturb.grouped, file = paste(OUTPUT_DIR, "sig_counts_perturb_grouped.txt", sep = "/"), row.names = FALSE, quote = FALSE)

# For each Hallmark gene set, how many significant perturbations are there?
hallmark.grouped <- analysis.df %>%
  group_by(pathway) %>%
  summarize(n_perturb_significant = sum(padj < 0.05, na.rm = TRUE)) %>%
  complete(pathway, fill = list(n_perturb_significant = 0))
write.table(hallmark.grouped, file = paste(OUTPUT_DIR, "sig_counts_hallmark_grouped.txt", sep = "/"), row.names = FALSE, quote = FALSE)
```

### Export the Analysis Matrix per Pathway too:

```{r}
```

### Creating a Correlation Matrix

```{r}
nt.df <- read.table("/raleighlab/data1/czou/gbm_perturb/gbm_perturb_gl261_clean_outputs/deseq/GL261_integrated_20230705_ced_noRTNormalized_all/CED_non-targeting_RT_non-targeting_noRT.csv")
View(nt.df
     )
```

```{r}
empty_plot <- ggplot() + theme_void()
combined.plot <- (empty_plot | plot_layout(ncol = 1, nrow = 1)) / 
                 (perturb.dendrogram | plot_layout(ncol = 1, nrow = 1)) / 
                 (module.dendrogram | bubble.plot + plot_layout(ncol = 1, nrow = 1)) / 
                 (empty_plot | plot_layout(ncol = 1, nrow = 1))
layout_matrix <- rbind(
  c(1, 2, 1),
  c(3, 4, NA),
  c(1, NA, NA)
)
combined.plot <- combined.plot + plot_layout(design = layout_matrix)
```

```{r}
perturb.dendrogram
```

```{r}
perturb.dendrogram + coord_flip() + scale_x_reverse()
```

```{r}
write.table(sgAnnot, paste(OUTPUT_DIR, "screen_phenotypes.txt", sep = "/"), sep = "\t")
```

```{r}
# Inspecting some GSEA results that didn't look quite right

ced.nt.table <- read.table("/raleighlab/data1/czou/gbm_perturb/gbm_perturb_gl261_clean_outputs/deseq/GL261_integrated_20230705_ced_noRTNormalized_all/CED_non-targeting_RT_non-targeting_noRT.csv")
ced.nt.table <- ced.nt.table[c("feature", "log_fc", "padj", "pvalue")]
ced.nt.table <- ced.nt.table[!is.na(ced.nt.table$pvalue), ]
ced.nt.table <- ced.nt.table[!is.na(ced.nt.table$padj), ]
colnames(ced.nt.table) <- c("feature", "ced_log_fc", "ced_padj", "ced_pvalue")

preinf.nt.table <- read.table("/raleighlab/data1/czou/gbm_perturb/gbm_perturb_gl261_clean_outputs/deseq/GL261_integrated_20230705_preinf_noRTNormalized_sorted/preinf_non-targeting_RT_non-targeting_noRT.csv")
preinf.nt.table <- preinf.nt.table[c("feature", "log_fc", "padj", "pvalue")]
preinf.nt.table <- preinf.nt.table[!is.na(preinf.nt.table$pvalue), ]
preinf.nt.table <- preinf.nt.table[!is.na(preinf.nt.table$padj), ]
colnames(preinf.nt.table) <- c("feature", "preinf_log_fc", "preinf_padj", "preinf_pvalue")

invitro.nt.table <- read.table("/raleighlab/data1/czou/gbm_perturb/gbm_perturb_gl261_clean_outputs/deseq/GL261_integrated_20230705_invitro_noRTNormalized_all/invitro_non-targeting_RT_non-targeting_noRT.csv")
invitro.nt.table <- invitro.nt.table[c("feature", "log_fc", "padj", "pvalue")]
invitro.nt.table <- invitro.nt.table[!is.na(invitro.nt.table$pvalue), ]
invitro.nt.table <- invitro.nt.table[!is.na(invitro.nt.table$padj), ]
colnames(invitro.nt.table) <- c("feature", "invitro_log_fc", "invitro_padj", "invitro_pvalue")

all.df <- merge(ced.nt.table, preinf.nt.table, by = "feature")
all.df <- merge(all.df, invitro.nt.table, by = "feature")
```
